### 2019-05-24 paper/paper.tex
abstract and first paragraphs

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}

\begin{document}

\input{header}

\maketitle

\begin{abstract}

Sorting is a fundamental programming problem with many important applications, being very useful to organize data for fast retrieval. While there are well-known sorting algorithms for conventional computers, especial techniques are required to take the most out of real hardware offering features such as cache memory and instruction-level parallelism. ChipSort is a Julia module for SIMD and cache-aware sorting. It implements sorting networks and bitonic merge networks with SIMD instructions, with configurable vector sizes. It also implements Comb-sort, which lends itself easily to vectorization and achieves good performance in arrays that fit cache memory, with low penalty for random access. This also depends on a parallel in-place matrix transposition implemented in ChipSort. Large arrays are approached with a multi-way Merge-sort. The implementation of ChipSort itself is of interest from a programming languages perspective due to its use of metaprogramming, and more specifically Julia's generated functions. This enables tuning the code generation to specific tasks and hardware at runtime, a feat owed to Julia. This article presents the implemented techniques as well as experiments that demonstrate the increased speed and the hardware adaptation capability of the module.

\end{abstract}

\section{Introduction}

Sorting has enjoyed an important place within computer programming topics for a long time. Apart from having great importance for practical applications, it also attracts much attention as a theoretically interesting problem in itself~\cite{DBLP:books/lib/Knuth98a}. While sorting is a well-understood problem in general, with a few classic algorithms available, achieving optimal performance for specific problems and architectures may require a careful algorithm selection rather than just tuning parameters \cite[Part II, Introduction]{DBLP:books/daglib/0023376}.

The running time of a computer program is governed by many factors, starting with algorithmic complexity and the base clock speed of the processor. The observed performance in real computers has been increasingly depending in processor features that are not taken into account by the simplest models, though. These features include cache memory and paralellism at the processor and instruction level. Cache memory has become popular in consumer processors in the past couple of decades, allowing programs to often attain higher performance than would be possible with a simpler memory architecture, as long as memory access patterns exhibit temporal and spatial locality~\cite[Chapter 3]{Drepper07whatevery}. Writing software that can better cope with paralellism has also become necessary to exploit the full potential offered by modern and future processors~\cite{wilson2018}.

The practical and theoretical interest in sorting, and the great relevance of cache memory and paralellism in modern computing naturally lead to research on sorting algorithms that take these factors into account.
chache sort simd \cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15,DBLP:journals/pvldb/BalkesenATO13}.

The \LaTeX{} document preparation system is a special version of the
\TeX; typesetting program where-in a collection of \TeX; commands
are added to \LaTeX{} to simplify typesetting. Importantly, it allows
the author to concentrate on the logical structure of the document
rather than its visual layout.\vskip 6pt
Moreover, \LaTeX{} provides a consistent and comprehensive document
preparation interface. There are simple-to-use commands for
generating a table of contents, lists of figures and/or tables, and indexes.
\LaTeX{} can automatically number list entries, equations, figures,
tables, and footnotes, as well as articles, sections, and subsections.
Using this numbering system, bibliographic citations, page
references, and cross references to any other numbered entity (e.g.
article, section, equation, figure, list entry, etc.) become quite simple
and straightforward. The use of \LaTeX{} document classes allows
a simple change of class to transform the appearance of your document.\vskip 6pt
\LaTeX{} is a powerful tool for managing long and complex documents.
In particular, partial processing enables long documents to
be produced article by article without losing sequential information.
The use of document classes allows a simple change of style
(or style option) to transform the appearance of your document.

\section{The JuliaCon Article Class}
\label{sec:documentclass}
%
The juliacon class file preserves the standard LATEX{} interface such
that any document that can be produced using the standard LATEX{}
article class can also be produced with the class file.\vskip 6pt
It is likely that the make up will change after file submission. For
this reason, we ask you to ignore details such as slightly long lines,
page stretching, or figures falling out of synchronization, as these
details can be dealt with at a later stage.\vskip 6pt
Use should be made of symbolic references (\verb \ref ) in order to
protect against late changes of order, etc.

\section{USING THE JuliaCon Article CLASS FILE}

If the file \verb juliacon.cls  is not already in the appropriate system directory
for \LaTeX{} files, either arrange for it to be put there or copy
it to your working directory. The \verb juliacon  document class is implemented
as a complete class, not a document style option. In order to
use the \verb juliacon  document class, replace \verb article  by \verb juliacon  in the
\verb \documentclass  command at the beginning of your document:
\vskip 6pt
\begin{centering}
    \verb \documentclass{article}  \end{centering}
\vskip 6pt
replace by
\vskip 6pt
 \verb \documentclass{juliacon}  \vskip 6pt
In general, the following standard document \verb style  options should
{ \itshape not} be used with the {\footnotesize \itshape article} class file:
\begin{enumerate}
\item[(1)] \verb 10pt,  \verb 11pt,  \verb 12pt   ? unavailable;
\item[(2)] \verb twoside  (no associated style file) ? \verb twoside  is the default;
\item[(3)] \verb fleqn, \verb leqno, \verb titlepage ? should not be used;
\end{enumerate}

\section{Additional Document Style Options}
\label{sec:additional_doc}
%
The following additional style option is available with the \verb juliacon  class file:
\vskip 6pt
Please place any additional command definitions at the very start of
the \LaTeX{} file, before the \verb \begin{document} . For example, user-defined
\verb \def  and \verb \newcommand   commands that define macros for
technical expressions should be placed here. Other author-defined
macros should be kept to a minimum.
\vskip 6pt
Commands that differ from the standard \LaTeX{} interface, or that
are provided in addition to the standard interface, are explained in
this guide. This guide is not a substitute for the \LaTeX{} manual itself.
Authors planning to submit their papers in \LaTeX{} are advised to use
\verb \juliacon.cls  as early as possible in the creation of their files.

%
%
%
%
\begin{table*}[t]
\tabcolsep22pt
\tbl{If necessary, the tables can be extended both columns.}{
\begin{tabular}{|l|l|c|c|}\hline
Label & \multicolumn{1}{c|}{Description}
& Number of Users &
Number of Queries\\\hline
Test 1 & Training Data &
\smash{\raise-7pt\hbox{70}} & 104\\
\cline{1-2}\cline{4-4}
Test 2 & Testing Data I & & 105\\\hline
Test 3 & Testing Data II & 30 & 119\\\hline
& Total & 100 & 328\\\hline
\end{tabular}}
\label{tab:symbols}
\begin{tabnote}
This is an example of table footnote.
\end{tabnote}
\end{table*}
% \begin{figure*}[t]
% \centerline{\includegraphics[width=11cm]{juliagraphs.png}}
% \caption{If necessary, the images can be extended both columns.}
%   \label{fig:sample_image}
% \end{figure*}

\section{Additional features}
\label{sec:additional_faci}
In addition to all the standard \LaTeX{} design elements, the \verb juliacon  class file includes the following features:
In general, once you have used the additional \verb juliacon.cls facilities
in your document, do not process it with a standard \LaTeX{} class
file.

\subsection{Titles, Author's Name, and Affiliation}
\label{subsub:title_auth}
The title of the article, author's name, and affiliation are used at the
beginning of the article (for the main title). These can be produced
using the following code:

\begin{verbatim}
\title{ This is an example of article title} }
\author{
   \large 1st Author \\[-3pt]
   \normalsize 1st author's affiliation  \\[-3pt]
    \normalsize 1st line of address \\[-3pt]
    \normalsize 2nd line of address \\[-3pt]
    \normalsize	1st author's email address \\[-3pt]
  \and
   \large 2nd Author \\[-3pt]
   \normalsize 2nd author's affiliation  \\[-3pt]
    \normalsize 1st line of address \\[-3pt]
    \normalsize 2nd line of address \\[-3pt]
    \normalsize	2nd author's email address \\[-3pt]
\and
   \large 3rd Author \\[-3pt]
   \normalsize 3rd author's affiliation  \\[-3pt]
    \normalsize 1st line of address \\[-3pt]
    \normalsize 2nd line of address \\[-3pt]
    \normalsize	3rd author's email address \\[-3pt]
}
\maketitle
\end{verbatim}

\subsection{Writing Julia code}

A special environment is already defined for Julia code,
built on top of \textit{listings} and \textit{jlcode}.

\begin{verbatim}
\begin{lstlisting}[language = Julia]
using Plots

x = -3.0:0.01:3.0
y = rand(length(x))
plot(x, y)
\end{lstlisting}
\end{verbatim}
\begin{lstlisting}[language = Julia]
using Plots

x = -3.0:0.01:3.0
y = rand(length(x))
plot(x, y)
\end{lstlisting}

\subsection{Abstracts, Key words, term etc...}
\label{subsub:abs_key_etc}

At the beginning of your article, the title should be generated
in the usual way using the \verb \maketitle  command. For genaral tem and keywords use
\verb \terms ,
\verb \keywords  commands respectively. The abstract should be enclosed
within an abstract environment, All these environment
can be produced using the following code:
\begin{verbatim}
\terms{Experimentation, Human Factors}

\keywords{Face animation, image-based modelling...}

\begin{abstract}
In this paper, we propose a new method for the
systematic determination of the model's base of
time varying delay system. This method based on
the construction of the classification data related
to the considered system. The number, the orders,
the time delay and the parameters of the local
models are generated automatically without any
knowledge about the full operating range of the
process. The parametric identification of the local
models is realized by a new recursive algorithm for
on line identification of systems with unknown time
delay. The proposed algorithm allows simultaneous
estimation of time delay and parameters of
discrete-time systems. The effectiveness of
the new method has been illustrated through
simulation.
\end{abstract}

\end{verbatim}

\section{Some guidelines}
\label{sec:some_guide}
The following notes may help you achieve the best effects with the
\verb juliacon  class file.

\subsection{Sections}
\label{subsub:sections}
\LaTeXe{}  provides four levels of section headings and they are all
defined in the \verb juliacon  class file:
\begin{itemize}
\item \verb \section   \item \verb \subsection  \item \verb \subsubsection  \item \verb \paragraph  \end{itemize}
Section headings are automatically converted to allcaps style.
\subsection{Lists}
\label{sec:lists}
%
The \verb juliacon   class file provides unnumbered lists using the
\verb unnumlist  environment for example,

\begin{unnumlist}
\item First unnumbered item which has no label and is indented from the
left margin.
\item Second unnumbered item.
\item Third unnumbered item.
\end{unnumlist}
The unnumbered list which has no label and is indented from the
left margin. was produced by:
\begin{verbatim}
\begin{unnumlist}
\item First unnumbered item...
\item Second unnumbered item...
\item Third unnumbered item...
\end{unnumlist}
\end{verbatim}

The \verb juliacon   class file also provides hyphen list using the
\verb itemize  environment for example,
\begin{itemize}
\item First unnumbered bulleted item which has no label and is indented
from the left margin.
\item Second unnumbered bulleted item.
\item Third unnumbered bulleted item which has no label and is indented
from the left margin.
\end{itemize}
was produced by:
\begin{verbatim}
\begin{itemize}
\item First item...
\item Second item...
\item Third item...
\end{itemize}
\end{verbatim}

Numbered list is also provided in acmtog class file using the
enumerate environment for example,
\begin{enumerate}
\item The attenuated and diluted stellar radiation.
\item Scattered radiation, and
\item Reradiation from other grains.
\end{enumerate}

was produced by:
\begin{verbatim}
\begin{enumerate}
\item The attenuated...
\item Scattered radiation, and...
\item Reradiation from other grains...
\end{enumerate}
\end{verbatim}
\subsection{Illustrations (or figures)}
\label{subsub:sec_Illus}
The \verb juliacon   class file will cope with most of the positioning of
your illustrations and you should not normally use the optional positional
qualifiers on the \verb figure   environment that would override
these decisions.
\vskip 6pt

%
\begin{figure}[t]
\centerline{\includegraphics[width=4cm]{juliagraphs.png}}
\caption{This is example of the image in a column.}
	\label{fig:sample_figure}
\end{figure}

The figure \ref{fig:sample_figure} is taken from the JuliaGraphs
organization \footnote{https://github.com/JuliaGraphs}.

Figure captions should be \emph{below} the figure itself, therefore the
\verb \caption  command should appear after the figure or space left for
an illustration. For example, Figure 1 is produced using the following
commands:

\begin{verbatim}
\begin{figure}
\centerline{\includegraphics[width=20pc]{Graphics.eps}}
\caption{An example of the testing process for a
binary tree. The globa null hypothesis is tested
first at level $\alpha$ (a), and the level of
individual variables is reached last (d). Note
that individual hypotheses can be tested at
level $\alpha/4$ and not $\alpha/8$ as one might
expect at first.}
\label{sample-figure_2}
\end{figure}
\end{verbatim}
Figures can be resized using first and second argument of
\verb \includegraphics   command. First argument is used for modifying
figure height and the second argument is used for modifying
figure width respectively.
\vskip 6pt
Cross-referencing of figures, tables, and numbered, displayed
equations using the \verb \label  and \verb \ref   commands is encouraged.
For example, in referencing Figure 1 above, we used
\verb Figure~\ref{sample-figure}   \subsection{Tables}
\label{subsub:sec_Tab}
The \verb juliacon   class file will cope with most of the positioning of
your tables and you should not normally use the optional positional qualifiers on the table environment which would override these
decisions. Table captions should be at the top.
\begin{verbatim}
\begin{table}
\tbl{Tuning Set and Testing Set}{
\begin{tabular}{|l|l|c|c|}\hline
Label & \multicolumn{1}{c|}{Description}
& Number of Users &
Number of Queries\\\hline
Train70 & Training Data &
\smash{\raise-7pt\hbox{70}} & 104\\
\cline{1-2}\cline{4-4}
Test70 & Testing Data I & & 105\\\hline
Test30 & Testing Data II & 30 & 119\\\hline
& Total & 100 & 328\\\hline
\end{tabular}}
\end{table}\end{verbatim}

\begin{table}
\tbl{Tuning Set and Testing Set}{
\begin{tabular}{|l|l|c|c|}\hline
Label & \multicolumn{1}{c|}{Description}
& Number of Users &
Number of Queries\\\hline
Test 1 & Training Data &
\smash{\raise-7pt\hbox{70}} & 104\\
\cline{1-2}\cline{4-4}
Test 2 & Testing Data I & & 105\\\hline
Test 3 & Testing Data II & 30 & 119\\\hline
& Total & 100 & 328\\\hline
\end{tabular}}
\end{table}
\subsection{Landscaping Pages}
\label{subsub:landscaping_pages}
If a table is too wide to fit the standard measure, it may be turned,
with its caption, to 90 degrees. Landscape tables cannot be produced
directly using the \verb juliacon   class file because \TeX{} itself cannot
turn the page, and not all device drivers provide such a facility.
The following procedure can be used to produce such pages.
\vskip 6pt
Use the package \verb rotating   in your document and change the coding
from
\begin{verbatim}
\begin{table}...\end{table}
to
\begin{sidewaystable}...\end{sidewaystable}
and for figures
\begin{figure}...\end{figure}
to
\begin{sidewaysfigure}...\end{sidewaysfigure}
\end{verbatim}

environments in your document to turn your table on the appropriate
page of your document. For instance, the following code prints
a page with the running head, a message half way down and the
table number towards the bottom.
\begin{verbatim}
\begin{sidewaystable}
\tbl{Landscape table caption to go here.}{...}
\label{landtab}
\end{sidewaystable}
\end{verbatim}

\subsection{Double Column Figure and Tables}
\label{subsub:double_fig_tab}
For generating the output of figures and tables in double column
we can use the following coding:

\begin{enumerate}
\item For Figures:
\begin{verbatim}
\begin{figure*}...\end{figure*}
\end{verbatim}
\item For landscape figures:
\begin{verbatim}
\begin{sidewaysfigure*}...\end{sidewaysfigure*}
\end{verbatim}
\item For Tables:
\begin{verbatim}
\begin{table*}...\end{table*}
\end{verbatim}
\item For landscape tables:
\begin{verbatim}
\begin{sidewaystable*}...\end{sidewaystable*}
\end{verbatim}
\end{enumerate}

\subsection{Typesetting Mathematics}
\label{subsub:type_math}
The \verb juliacon class file will set displayed mathematics with center to
the column width, provided that you use the \LaTeXe{} standard of
open and closed square brackets as delimiters.
The equation
\[
\sum_{i=1}^p \lambda_i = (S)
\]

was typeset using the acmtog class file with the commands

\begin{verbatim}
\[
\sum_{i=1}^p \lambda_i = (S)
\]
\end{verbatim}

For display equations, cross-referencing is encouraged. For example,
\begin{verbatim}
\begin{equation}
(n-1)^{-1} \sum^n_{i=1} (X_i - \overline{X})^2.
\label{eq:samplevar}
\end{equation}
Equation~(\ref{eq:samplevar}) gives the formula for
sample variance.
\end{verbatim}
The following output is generated with the above coding:
\begin{equation}
(n-1)^{-1} \sum^n_{i=1} (X_i - \overline{X})^2.
\label{eq:samplevar}
\end{equation}
Equation~(\ref{eq:samplevar}) gives the formula for
sample variance.

\subsection{Enunciations}
\label{subsub:enunciation}
The \verb juliacon   class file generates the enunciations with the help of
the following commands:
\begin{verbatim}
\begin{theorem}...\end{theorem}
\begin{strategy}...\end{strategy}
\begin{property}...\end{property}
\begin{proposition}...\end{proposition}
\begin{lemma}...\end{lemma}
\begin{example}...\end{example}
\begin{proof}...\end{proof}
\begin{definition}...\end{definition}
\begin{algorithm}...\end{algorithm}
\begin{remark}...\end{remark}
\end{verbatim}
The above-mentioned coding can also include optional arguments
such as
\begin{verbatim}
\begin{theorem}[...]. Example for theorem:
\begin{theorem}[Generalized Poincare Conjecture]
Four score and seven ... created equal.
\end{theorem}
\end{verbatim}

\begin{theorem}[Generalized Poincare Conjecture]
Four score and seven years ago our fathers brought forth,
upon this continent, a new nation, conceived in Liberty,
 and dedicated to the proposition that all men are
created equal.
\end{theorem}

\subsection{Extract}
\label{subsub:extract}
Extract environment should be coded within
\begin{verbatim}
\begin{extract}..\end{extract}
\end{verbatim}

\subsection{Balancing column at last page}
\label{subsub:Balance}
For balancing the both column length at last page use :
\begin{verbatim}
\vadjust{\vfill\pagebreak}
\end{verbatim}

%\vadjust{\vfill\pagebreak}

at appropriate place in your \TeX{} file or in bibliography file.

\section{Handling references}
\label{subsub:references}
References are most easily (and correctly) generated using the
BIBTEX, which is easily invoked via
\begin{verbatim}
\bibliographystyle{juliacon}
\bibliography{ref}
\end{verbatim}
When submitting the document source (.tex) file to external
parties, the ref.bib file should be sent with it.
\cite{bezanson2017julia}

\input{bib.tex}

\end{document}

% Inspired by the International Journal of Computer Applications template

### 2019-05-25 paper/paper.tex
finishing sec 1

@@
 
-The practical and theoretical interest in sorting, and the great relevance of cache memory and paralellism in modern computing naturally lead to research on sorting algorithms that take these factors into account.
-chache sort simd \cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15,DBLP:journals/pvldb/BalkesenATO13}.
+The interest in sorting and the great relevance of cache memory and paralellism in modern computing naturally lead to research on sorting algorithms that account for these factors.
+Processors offering SIMD (single instruction multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
+One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}.
+The current availability of processors with 512 bit registers underscore the need of being able to write software that exploit vector operations, not to mention GPUs.
+
+At the time CPUs with SIMD instructions became widely available there was already a trend to employ specialized hardware like GPUs for scientific applications~\cite{larsen2001fast,DBLP:conf/micro/ThompsonHO02}.
+Althoug SIMD was used mostly for multimedia applications at first~\cite{CHEN2006509,DBLP:journals/mam/SlingerlandS05}, we can cite some early alternative uses including 3D graphics~\cite{DBLP:conf/pcm/MaY02} and machine learning~\cite{DBLP:conf/europar/StreyB01}.
+
+Apart from earlier work on parallel sorting in other architectures, the late 2000s saw the first practical demonstrations of sorting with modern consumer SIMD chips~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
+Most of the techniques utilized then are still employed in more recent works~\cite{DBLP:journals/pvldb/BalkesenATO13,DBLP:journals/pvldb/InoueT15}.
+
+This article presents \verb ChipSort, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
+These include sorting and merging networks, the use of Comb-sort for medium sized arrays and of multi-way merging for large arrays.
+These techniques require not only tuning parameters such as vector and array sizes, but also generating specific code when implementing different size networks.
+\verb ChipSort seeks to generate efficient custom code for each task, relying on Julia metaprogramming capabilities for that.
+
+Following Section~\ref{sec:methods} presents the techniques employed in ChipSort.
+Section~\ref{sec:experiments} presents experimental results assessing the performance of the module, and Section~\ref{sec:conclusion} brings some concluding remarks.
+
+\section{SIMD sorting methods}
+\label{sec:methods}
+%
+\subsection{Sorting Networks}
+%
+\subsection{Bitonic Merge Networks}
+%
+\subsection{Vectorized Comb-sort}
+%
+\subsection{In-place matrix transpose}
+%
+\subsection{Vectorized multi-way array merging}
+%
+
+\section{Experiments}
+\label{sec:experiments}
+%
+\section{Conclusion}
+\label{sec:conclusion}
+%
 

### 2019-05-25 paper/paper.tex
sorting networks

@@
 Following Section~\ref{sec:methods} presents the techniques employed in ChipSort.
-Section~\ref{sec:experiments} presents experimental results assessing the performance of the module, and Section~\ref{sec:conclusion} brings some concluding remarks.
+Section~\ref{sec:experiments} presents experimental results assessing the performance of the module and Section~\ref{sec:conclusion} brings some concluding remarks.
 
@@
 %
+ChipSort offers three high-level functions that can be used in applications as replacements for other sorting functions such as \verb sort  in the Julia standard library.
+Additional work is required from the user selecting among the three functions and also setting some numeric parameters.
+The functions are intended to be used with arrays of different sizes, and \verb chipsort_large  offers a good overview of all the techniques implemented in the module.
+Some of these techniques might also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.
+ChipSort does not currently support registers with separate key and payload, what might allow significant optimizations.
+Only the sorting of simple data types such as 32 and 64-bit integers and floating-point numbers were studied.
+
+This section will present how \verb chipsort_large  works, detailing each step of the process.
+It mostly follows~\cite{DBLP:journals/pvldb/InoueT15} with a few modifications.
+This function achieves sorting by first splitting the input array in chunks that are sorted separately with \verb chipsort_medium!
+and then merged simultaneously.
+This inner sort is performed in multiple steps:
+\begin{enumerate}
+\item sorting small blocks of data with a sorting network;
+\item reordering vectors in-place according to a matrix transpose;
+\item vectorized Comb-sort with a limited number of iterations;
+\item regular insertion sort until the whole data is sorted.
+\end{enumerate}  
+The \verb chipsort_small!  function is essentially a sorting network followed by a merge procedure.
+
+The remainder of this section will detail the multiple techniques employed in this process.
+
 \subsection{Sorting Networks}
 %
+Sorting networks~\cite[Sec.
+5.3.4]{DBLP:books/lib/Knuth98a} are graphs composed by comparator modules, units that take two numbers in, and output them in order: the smallest and the largest at specific outputs.
+Properly arranging the compatators allows the creation of a procedure that takes a vector of numbers in, and outputs them in sequence.
+The design of a sorting network can minimize the total number of comparisons, or also take into account that comparisons can be done in parallel, minimizing the number of steps to complete the process.
+
+A sorting network can be represented by a Knuth graph such as Figure~\ref{fig:sorting-network}.
+The graph represents a network that takes an array of four values $(A_1, A_2, A_3, A_4)$ and outputs the sorted sequence $(S_1, S_2, S_3, S_4)$.
+Each vertical line in the graph represents a comparator.
+\begin{figure}[htb]
+\centerline{\includegraphics[width=0.9\linewidth]{fig/sorting-network-4.pdf}}
+\caption{A sorting network for 4 elements. The dashed lines represent sections where operations are independent of each other.}
+\label{fig:sorting-network}
+\end{figure}
+
+In ChipSort, as in the related work, parallelism is mainly exploited by carrying out the comparisons of the sorting network over vectors of size $V$ contained in SIMD registers, effectively running the network $V$ times simultaneously.
+Other forms of instruction-level parallelism can also be in play, though.
+In especial, it may also be possible to parallelize instructions such as \verb min  and \verb max , and also the loading of data into registers.
+This is often done implicitly by the microarchitecture, and programmers can only try to make sure their code is suitable for this by properly ordering operations and avoiding branching, for instance.
+Part of this work is expected to be performed by the compiler, and especially LLVM in the case of Julia.
+
+ChipSort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/ChipSort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}.
+The generated function \href{https://github.com/nlw0/ChipSort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} works by assigning each element of the sequence at each step to a variable, and calculating values from the next step according to the network specification.
+The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types.
+Figure~\ref{fig:sort-net-listing}
+
+\begin{figure}[htb]
+\begin{lstlisting}[language = Julia]
+input_0_1 = Base.getindex(input, 1)
+input_0_2 = Base.getindex(input, 2)
+input_0_3 = Base.getindex(input, 3)
+input_0_4 = Base.getindex(input, 4)
+input_1_1 = ChipSort.min(input_0_1, input_0_2)
+input_1_2 = ChipSort.max(input_0_1, input_0_2)
+input_1_3 = ChipSort.min(input_0_3, input_0_4)
+input_1_4 = ChipSort.max(input_0_3, input_0_4)
+input_2_1 = ChipSort.min(input_1_1, input_1_3)
+input_2_3 = ChipSort.max(input_1_1, input_1_3)
+input_2_2 = ChipSort.min(input_1_2, input_1_4)
+input_2_4 = ChipSort.max(input_1_2, input_1_4)
+input_3_1 = input_2_1
+input_3_4 = input_2_4
+input_3_2 = ChipSort.min(input_2_2, input_2_3)
+input_3_3 = ChipSort.max(input_2_2, input_2_3)
+return (input_3_1, input_3_2, input_3_3, input_3_4)
+\end{lstlisting}
+\caption{Code generated by {\tt sort\_net} to implement a 4-element sorting network.}
+\label{fig:sort-net-listing}
+\end{figure}
+
+This function is a good first example of the utilization of Julia's meta-programming features in ChipSort.
+Multiple implementations of sorting networks are available~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08,sortingnetworksjl,ultrasort}, however they implement networks of different sizes straight as code.
+In ChipSort they are represented as data structures that guide a generated function that produces equivalent code.
+While this data structure currently is hard-coded, it can potentially be produced only when necessary by an implementation of an algorithm such as Bose-Nelson, what remains a future plan for the module.
+
+The {\tt sort\_net} generated function expects elements that have defined {\tt max } and {\tt min} methods.
+It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works.
+In practice ChipSort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}.
+This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible.
+
+\subsection{SIMD}
+
+When applied to a group of vectors, the result from the sorting network contains sorted sequences in the lanes of the vectors. To obtain sorted vectors we need to transpose these vectors.
+
 \subsection{Bitonic Merge Networks}
@@
 % Inspired by the International Journal of Computer Applications template
+
+% latexmk -bibtex -pdf paper.tex

### 2019-05-25 paper/paper.tex
vectors transpose

@@
 
-\section{SIMD sorting methods}
+\section{SIMD sorting techniques}
 \label{sec:methods}
@@
 \begin{figure}[htb]
-\centerline{\includegraphics[width=0.9\linewidth]{fig/sorting-network-4.pdf}}
-\caption{A sorting network for 4 elements. The dashed lines represent sections where operations are independent of each other.}
+\centerline{\includegraphics[width=0.6\linewidth]{fig/sorting-network-4.pdf}}
+\caption{A sorting network for 4 elements. The dashed lines delimit each step where operations are independent of each other.}
 \label{fig:sorting-network}
@@
 \begin{lstlisting}[language = Julia]
-input_0_1 = Base.getindex(input, 1)
-input_0_2 = Base.getindex(input, 2)
-input_0_3 = Base.getindex(input, 3)
-input_0_4 = Base.getindex(input, 4)
-input_1_1 = ChipSort.min(input_0_1, input_0_2)
-input_1_2 = ChipSort.max(input_0_1, input_0_2)
-input_1_3 = ChipSort.min(input_0_3, input_0_4)
-input_1_4 = ChipSort.max(input_0_3, input_0_4)
-input_2_1 = ChipSort.min(input_1_1, input_1_3)
-input_2_3 = ChipSort.max(input_1_1, input_1_3)
-input_2_2 = ChipSort.min(input_1_2, input_1_4)
-input_2_4 = ChipSort.max(input_1_2, input_1_4)
+4 => (((1,2), (3,4)), ((1,3), (2,4)), ((2,3),))
+\end{lstlisting}
+\begin{lstlisting}[language = Julia]
+input_0_1 = input[1]
+input_0_2 = input[2]
+input_0_3 = input[3]
+input_0_4 = input[4]
+input_1_1 = min(input_0_1, input_0_2)
+input_1_2 = max(input_0_1, input_0_2)
+input_1_3 = min(input_0_3, input_0_4)
+input_1_4 = max(input_0_3, input_0_4)
+input_2_1 = min(input_1_1, input_1_3)
+input_2_3 = max(input_1_1, input_1_3)
+input_2_2 = min(input_1_2, input_1_4)
+input_2_4 = max(input_1_2, input_1_4)
 input_3_1 = input_2_1
 input_3_4 = input_2_4
-input_3_2 = ChipSort.min(input_2_2, input_2_3)
-input_3_3 = ChipSort.max(input_2_2, input_2_3)
+input_3_2 = min(input_2_2, input_2_3)
+input_3_3 = max(input_2_2, input_2_3)
 return (input_3_1, input_3_2, input_3_3, input_3_4)
 \end{lstlisting}
-\caption{Code generated by {\tt sort\_net} to implement a 4-element sorting network.}
+\caption{The parameters from a four-elements sorting network, and the corresponding code generated by {\tt sort\_net} to implement it.}
 \label{fig:sort-net-listing}
@@
 
-\subsection{SIMD}
+\subsection{SIMD vectors transpose}
 
-When applied to a group of vectors, the result from the sorting network contains sorted sequences in the lanes of the vectors. To obtain sorted vectors we need to transpose these vectors.
+Consider a group of stacked SIMD vectors.
+Each column in this representation is termed a {\em lane}.
+The result from {\tt sort\_vecs} is that each lane contains a sorted sequence.
+To obtain sorted vectors we need to transpose these stacked vectors.
+This is done in ChipSort through {\tt transpose\_vecs}, which is a generated function that supports rectangular matrices, but with dimensions constrained to powers of 2.
+This function is heavily based on calls to {\tt SIMD.shufflevec}.
+
+Figure~\ref{fig:transpose-vecs} displays an example of 64 random integers loaded into SIMD registers as 4 vectors.
+After {\tt sort\_vecs} each lane contains a sorted sequence.
+After {\tt transpose\_vecs} each vector (register) contains a sorted sequence.
+
+\begin{figure}[htb]
+\centerline{\includegraphics[width=0.99\linewidth]{fig/regs.pdf}}
+\caption{Four vectors with four values, the result from a vectorized four-elements sorting network, and the transpose.}
+\label{fig:transpose-vecs}
+\end{figure}
 

### 2019-05-26 paper/paper.tex
transpose

@@
 It implements sorting networks and bitonic merge networks with SIMD instructions, with configurable vector sizes.
-It also implements Comb-sort, which lends itself easily to vectorization and achieves good performance in arrays that fit cache memory, with low penalty for random access.
+It also implements Combsort, which lends itself easily to vectorization and achieves good performance in arrays that fit cache memory, with low penalty for random access.
 This also depends on a parallel in-place matrix transposition implemented in ChipSort.
-Large arrays are approached with a multi-way Merge-sort.
+Large arrays are approached with a multi-way Mergesort.
 The implementation of ChipSort itself is of interest from a programming languages perspective due to its use of metaprogramming, and more specifically Julia's generated functions.
@@
 This article presents \verb ChipSort, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
-These include sorting and merging networks, the use of Comb-sort for medium sized arrays and of multi-way merging for large arrays.
+These include sorting and merging networks, the use of Combsort for medium sized arrays and of multi-way merging for large arrays.
 These techniques require not only tuning parameters such as vector and array sizes, but also generating specific code when implementing different size networks.
@@
 \item reordering vectors in-place according to a matrix transpose;
-\item vectorized Comb-sort with a limited number of iterations;
+\item vectorized Combsort with a limited number of iterations;
 \item regular insertion sort until the whole data is sorted.
 \end{enumerate}
-The \verb chipsort_small!  function is essentially a sorting network followed by a merge procedure.
 
@@
 \end{lstlisting}
-\caption{The parameters from a four-elements sorting network, and the corresponding code generated by {\tt sort\_net} to implement it.}
+\caption{The parameters from a four-element sorting network consisting of three steps, and the corresponding code generated by {\tt sort\_net}.}
 \label{fig:sort-net-listing}
@@
 
-The {\tt sort\_net} generated function expects elements that have defined {\tt max } and {\tt min} methods.
+The {\tt sort\_net} generated function expects elements with {\tt max} and {\tt min} methods defined.
 It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works.
 In practice ChipSort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}.
-This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible.
+This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers.
 
 \subsection{SIMD vectors transpose}
-
+%
 Consider a group of stacked SIMD vectors.
@@
 The result from {\tt sort\_vecs} is that each lane contains a sorted sequence.
-To obtain sorted vectors we need to transpose these stacked vectors.
+This data must be transposed to produce sorted vectors.
 This is done in ChipSort through {\tt transpose\_vecs}, which is a generated function that supports rectangular matrices, but with dimensions constrained to powers of 2.
-This function is heavily based on calls to {\tt SIMD.shufflevec}.
+The generated code consists mostly of calls to {\tt SIMD.shufflevec}.
+This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions, contemplating each different case, and in ChipSort there is only a single higher-level Julia generated function that produces equivalent code.
 
 Figure~\ref{fig:transpose-vecs} displays an example of 64 random integers loaded into SIMD registers as 4 vectors.
-After {\tt sort\_vecs} each lane contains a sorted sequence.
-After {\tt transpose\_vecs} each vector (register) contains a sorted sequence.
+After {\tt sort\_vecs} each lane contains a sorted sequence, and {\tt transpose\_vecs} turns the lanes into vectors (registers).
+The \verb chipsort_small!
+function is essentially a sorting network followed by a transpose and a merge procedure.
 
@@
 %
-\subsection{Vectorized Comb-sort}
+A bitonic sequence starts either increasing or decreasing and contains at most one change of direction.
+Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in ChipSort in a single generated function, {\tt bitonic\_merge}.
+Merging two sorted vectors requires first reversing one of them, and the result is stored in two new vectors of the same size containing the first and second halves of the total sorted sequence.
+
+Apart from the use in sorting small sequences and in the multi-way merge tree, bitonic merge networks can also be used to implement a regular merge sort.
+This technique was investigated in ChipSort only superficially, and while there were no promising results in terms of performance, ChipSort also contains the generated function {\tt bitonic\_merge\_interleaved} to illustrate how metaprogramming can help implementing multiple simultaneous merges with interleaved execution, as done in other projects~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
+
+\subsection{Vectorized Combsort}
 %
+One of the most peculiar aspects of ChipSort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm.
+Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes insertion sort~\cite{dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}.
+The array is sweeped multiple times, and pairs of values at a given distance from each other are compared and stored in order.
+The distance is reduced at each pass by a factor of $\frac{4}{3}$.
+This algorithm lends itself easily to vectorization: performing the comparisons over vectors of size $V$ effectively allows us to run the algorithm over $V$ lists simultaneouly.
+
+In practice, ChipSort performs a number of other operations before and after using Combsort. The {\tt chipsort\_medium!} function consists of the following steps:
+
+\begin{enumerate}
+\item Apply a sorting network on $K$ blocks of $J$ vectors of size $V$.
+\item Vectorized Combsort until the interval size is 1.
+\item Transpose blocks.
+\item Transpose data in-place from $K$ blocks of $J$ vectors into $J$ blocks of $K$ vectors.
+\item Vectorized Combsort again.
+\item Sort blocks again.
+\item Insertion sort.
+\end{enumerate}
+
+After the first two steps the result is essentially a matrix where the $K\times J$ columns are the vectors, and each row contains an approximately sorted sequence.
+The transpose steps reshape each row into a new block.
+Each row is processed independently at first, and the transpose allows elements from distinct groups to interact.
+
+The last step is necessary to ensure the whole array is sorted, not vieved anymore as a matrix with independent rows.
+Insertion sort also presents a better performance than sticking to Combsort until the end, what would essentially mean using bubble sort once the interval becomes 1.
+Because the array is now approximately sorted, it is expected at this time that insertion sort does not have to move any out-of-order elements very far from where they are found.
+
 \subsection{In-place matrix transpose}
 %
+In-place matrix transposition can be attained by moving a number to its destination, then moving the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}.
+While carrying out this procedure is simple, it is not trivial to find out which are the cycles to tranpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[Sec.
+1.3.3, Ex.
+12]{DBLP:books/lib/Knuth97}.
+
+ChipSort contains the generated function {\tt transpose!} to perform in-place matrix transposition.
+The cycle seeds are computed at the time of code generation, and at runtime the function only needs to move the data and compute the indices detecting when each cycle finishes.
+The function is therefore completely generic, supporting any dimensions, but with a very simple runtime code, with the most complex part of the problem being tackled programatically during code generation.
+
 \subsection{Vectorized multi-way array merging}

### 2019-05-26 paper/paper.tex
multiwaymerge

@@
 
-\subsection{Sorting Networks}
+\subsection{Sorting networks}
 %
@@
 
-\subsection{Vectorized multi-way array merging}
+\subsection{Vectorized multi-way merging}
 %
+Merging multiple arrays simultaneously requires the creation of a {\em merge tree} to keep some intermediate data.
+Each node in this tree is essentially an iterator traversing the merger of two arrays in the lower level.
+To employ the vectorized bitonic merge network in this process it is necessary to keep a whole vector at each node of the tree.
+To produce the next vector of elements one of the two source arrays is selected according to their smallest next element, and a whole vector is pulled from that array.
+This vector is merged with the intermediate one, and the first half of the resulting sequence can be taken to the next level of the tree, while the second half is kept as the new intermediate data.
+
+Given a set of sorted arrays, once the tree is initialized data is taken from it one vector at a time, producing the final sorted sequence.
+Each time a new vector is requested the tree is traversed looking for the smallest element between each of the node children, until one of the input arrays is reached.
+The data from the array is then loaded from memory and merged with the intermediate data, producing new updates moving trought the tree until the root is reached.
+
+The intermediate data from the tree is kept by ChipSort in a contiguous array, treated as a heap.
+Each node ancestor can be found simply by the Euclidean division of its index by 2.
+During operation it is assumed this data should fit some level of cache memory, as in the related work~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15}.
 
@@
 
-The \LaTeX{} document preparation system is a special version of the
-\TeX; typesetting program where-in a collection of \TeX; commands
-are added to \LaTeX{} to simplify typesetting. Importantly, it allows
-the author to concentrate on the logical structure of the document
-rather than its visual layout.\vskip 6pt
-Moreover, \LaTeX{} provides a consistent and comprehensive document
-preparation interface. There are simple-to-use commands for
-generating a table of contents, lists of figures and/or tables, and indexes.
-\LaTeX{} can automatically number list entries, equations, figures,
-tables, and footnotes, as well as articles, sections, and subsections.
-Using this numbering system, bibliographic citations, page
-references, and cross references to any other numbered entity (e.g.
-article, section, equation, figure, list entry, etc.) become quite simple
-and straightforward. The use of \LaTeX{} document classes allows
-a simple change of class to transform the appearance of your document.\vskip 6pt
-\LaTeX{} is a powerful tool for managing long and complex documents.
-In particular, partial processing enables long documents to
-be produced article by article without losing sequential information.
-The use of document classes allows a simple change of style
-(or style option) to transform the appearance of your document.
-
-\section{The JuliaCon Article Class}
-\label{sec:documentclass}
-%
-The juliacon class file preserves the standard LATEX{} interface such
-that any document that can be produced using the standard LATEX{}
-article class can also be produced with the class file.\vskip 6pt
-It is likely that the make up will change after file submission. For
-this reason, we ask you to ignore details such as slightly long lines,
-page stretching, or figures falling out of synchronization, as these
-details can be dealt with at a later stage.\vskip 6pt
-Use should be made of symbolic references (\verb \ref ) in order to
-protect against late changes of order, etc.
-
-\section{USING THE JuliaCon Article CLASS FILE}
-
-If the file \verb juliacon.cls  is not already in the appropriate system directory
-for \LaTeX{} files, either arrange for it to be put there or copy
-it to your working directory. The \verb juliacon  document class is implemented
-as a complete class, not a document style option. In order to
-use the \verb juliacon  document class, replace \verb article  by \verb juliacon  in the
-\verb \documentclass  command at the beginning of your document:
-\vskip 6pt
-\begin{centering}
-    \verb \documentclass{article}  \end{centering}
-\vskip 6pt
-replace by
-\vskip 6pt
- \verb \documentclass{juliacon}  \vskip 6pt
-In general, the following standard document \verb style  options should
-{ \itshape not} be used with the {\footnotesize \itshape article} class file:
-\begin{enumerate}
-\item[(1)] \verb 10pt,  \verb 11pt,  \verb 12pt   ? unavailable;
-\item[(2)] \verb twoside  (no associated style file) ? \verb twoside  is the default;
-\item[(3)] \verb fleqn, \verb leqno, \verb titlepage ? should not be used;
-\end{enumerate}
-
-\section{Additional Document Style Options}
-\label{sec:additional_doc}
-%
-The following additional style option is available with the \verb juliacon  class file:
-\vskip 6pt
-Please place any additional command definitions at the very start of
-the \LaTeX{} file, before the \verb \begin{document} . For example, user-defined
-\verb \def  and \verb \newcommand   commands that define macros for
-technical expressions should be placed here. Other author-defined
-macros should be kept to a minimum.
-\vskip 6pt
-Commands that differ from the standard \LaTeX{} interface, or that
-are provided in addition to the standard interface, are explained in
-this guide. This guide is not a substitute for the \LaTeX{} manual itself.
-Authors planning to submit their papers in \LaTeX{} are advised to use
-\verb \juliacon.cls  as early as possible in the creation of their files.
-
-%
-%
-%
-%
-\begin{table*}[t]
-\tabcolsep22pt
-\tbl{If necessary, the tables can be extended both columns.}{
-\begin{tabular}{|l|l|c|c|}\hline
-Label & \multicolumn{1}{c|}{Description}
-& Number of Users &
-Number of Queries\\\hline
-Test 1 & Training Data &
-\smash{\raise-7pt\hbox{70}} & 104\\
-\cline{1-2}\cline{4-4}
-Test 2 & Testing Data I & & 105\\\hline
-Test 3 & Testing Data II & 30 & 119\\\hline
-& Total & 100 & 328\\\hline
-\end{tabular}}
-\label{tab:symbols}
-\begin{tabnote}
-This is an example of table footnote.
-\end{tabnote}
-\end{table*}
-% \begin{figure*}[t]
-% \centerline{\includegraphics[width=11cm]{juliagraphs.png}}
-% \caption{If necessary, the images can be extended both columns.}
-%   \label{fig:sample_image}
-% \end{figure*}
-
-\section{Additional features}
-\label{sec:additional_faci}
-In addition to all the standard \LaTeX{} design elements, the \verb juliacon  class file includes the following features:
-In general, once you have used the additional \verb juliacon.cls facilities
-in your document, do not process it with a standard \LaTeX{} class
-file.
-
-\subsection{Titles, Author's Name, and Affiliation}
-\label{subsub:title_auth}
-The title of the article, author's name, and affiliation are used at the
-beginning of the article (for the main title). These can be produced
-using the following code:
-
-\begin{verbatim}
-\title{ This is an example of article title} }
-\author{
-   \large 1st Author \\[-3pt]
-   \normalsize 1st author's affiliation  \\[-3pt]
-    \normalsize 1st line of address \\[-3pt]
-    \normalsize 2nd line of address \\[-3pt]
-    \normalsize	1st author's email address \\[-3pt]
-  \and
-   \large 2nd Author \\[-3pt]
-   \normalsize 2nd author's affiliation  \\[-3pt]
-    \normalsize 1st line of address \\[-3pt]
-    \normalsize 2nd line of address \\[-3pt]
-    \normalsize	2nd author's email address \\[-3pt]
-\and
-   \large 3rd Author \\[-3pt]
-   \normalsize 3rd author's affiliation  \\[-3pt]
-    \normalsize 1st line of address \\[-3pt]
-    \normalsize 2nd line of address \\[-3pt]
-    \normalsize	3rd author's email address \\[-3pt]
-}
-\maketitle
-\end{verbatim}
-
-\subsection{Writing Julia code}
-
-A special environment is already defined for Julia code,
-built on top of \textit{listings} and \textit{jlcode}.
-
-\begin{verbatim}
-\begin{lstlisting}[language = Julia]
-using Plots
-
-x = -3.0:0.01:3.0
-y = rand(length(x))
-plot(x, y)
-\end{lstlisting}
-\end{verbatim}
-\begin{lstlisting}[language = Julia]
-using Plots
-
-x = -3.0:0.01:3.0
-y = rand(length(x))
-plot(x, y)
-\end{lstlisting}
-
-\subsection{Abstracts, Key words, term etc...}
-\label{subsub:abs_key_etc}
-
-At the beginning of your article, the title should be generated
-in the usual way using the \verb \maketitle  command. For genaral tem and keywords use
-\verb \terms ,
-\verb \keywords  commands respectively. The abstract should be enclosed
-within an abstract environment, All these environment
-can be produced using the following code:
-\begin{verbatim}
-\terms{Experimentation, Human Factors}
-
-\keywords{Face animation, image-based modelling...}
-
-\begin{abstract}
-In this paper, we propose a new method for the
-systematic determination of the model's base of
-time varying delay system. This method based on
-the construction of the classification data related
-to the considered system. The number, the orders,
-the time delay and the parameters of the local
-models are generated automatically without any
-knowledge about the full operating range of the
-process. The parametric identification of the local
-models is realized by a new recursive algorithm for
-on line identification of systems with unknown time
-delay. The proposed algorithm allows simultaneous
-estimation of time delay and parameters of
-discrete-time systems. The effectiveness of
-the new method has been illustrated through
-simulation.
-\end{abstract}
-
-\end{verbatim}
-
-\section{Some guidelines}
-\label{sec:some_guide}
-The following notes may help you achieve the best effects with the
-\verb juliacon  class file.
-
-\subsection{Sections}
-\label{subsub:sections}
-\LaTeXe{}  provides four levels of section headings and they are all
-defined in the \verb juliacon  class file:
-\begin{itemize}
-\item \verb \section   \item \verb \subsection  \item \verb \subsubsection  \item \verb \paragraph  \end{itemize}
-Section headings are automatically converted to allcaps style.
-\subsection{Lists}
-\label{sec:lists}
-%
-The \verb juliacon   class file provides unnumbered lists using the
-\verb unnumlist  environment for example,
-
-\begin{unnumlist}
-\item First unnumbered item which has no label and is indented from the
-left margin.
-\item Second unnumbered item.
-\item Third unnumbered item.
-\end{unnumlist}
-The unnumbered list which has no label and is indented from the
-left margin. was produced by:
-\begin{verbatim}
-\begin{unnumlist}
-\item First unnumbered item...
-\item Second unnumbered item...
-\item Third unnumbered item...
-\end{unnumlist}
-\end{verbatim}
-
-The \verb juliacon   class file also provides hyphen list using the
-\verb itemize  environment for example,
-\begin{itemize}
-\item First unnumbered bulleted item which has no label and is indented
-from the left margin.
-\item Second unnumbered bulleted item.
-\item Third unnumbered bulleted item which has no label and is indented
-from the left margin.
-\end{itemize}
-was produced by:
-\begin{verbatim}
-\begin{itemize}
-\item First item...
-\item Second item...
-\item Third item...
-\end{itemize}
-\end{verbatim}
-
-Numbered list is also provided in acmtog class file using the
-enumerate environment for example,
-\begin{enumerate}
-\item The attenuated and diluted stellar radiation.
-\item Scattered radiation, and
-\item Reradiation from other grains.
-\end{enumerate}
-
-was produced by:
-\begin{verbatim}
-\begin{enumerate}
-\item The attenuated...
-\item Scattered radiation, and...
-\item Reradiation from other grains...
-\end{enumerate}
-\end{verbatim}
-\subsection{Illustrations (or figures)}
-\label{subsub:sec_Illus}
-The \verb juliacon   class file will cope with most of the positioning of
-your illustrations and you should not normally use the optional positional
-qualifiers on the \verb figure   environment that would override
-these decisions.
-\vskip 6pt
-
-%
-\begin{figure}[t]
-\centerline{\includegraphics[width=4cm]{juliagraphs.png}}
-\caption{This is example of the image in a column.}
-	\label{fig:sample_figure}
-\end{figure}
-
-The figure \ref{fig:sample_figure} is taken from the JuliaGraphs
-organization \footnote{https://github.com/JuliaGraphs}.
-
-Figure captions should be \emph{below} the figure itself, therefore the
-\verb \caption  command should appear after the figure or space left for
-an illustration. For example, Figure 1 is produced using the following
-commands:
-
-\begin{verbatim}
-\begin{figure}
-\centerline{\includegraphics[width=20pc]{Graphics.eps}}
-\caption{An example of the testing process for a
-binary tree. The globa null hypothesis is tested
-first at level $\alpha$ (a), and the level of
-individual variables is reached last (d). Note
-that individual hypotheses can be tested at
-level $\alpha/4$ and not $\alpha/8$ as one might
-expect at first.}
-\label{sample-figure_2}
-\end{figure}
-\end{verbatim}
-Figures can be resized using first and second argument of
-\verb \includegraphics   command. First argument is used for modifying
-figure height and the second argument is used for modifying
-figure width respectively.
-\vskip 6pt
-Cross-referencing of figures, tables, and numbered, displayed
-equations using the \verb \label  and \verb \ref   commands is encouraged.
-For example, in referencing Figure 1 above, we used
-\verb Figure~\ref{sample-figure}   \subsection{Tables}
-\label{subsub:sec_Tab}
-The \verb juliacon   class file will cope with most of the positioning of
-your tables and you should not normally use the optional positional qualifiers on the table environment which would override these
-decisions. Table captions should be at the top.
-\begin{verbatim}
-\begin{table}
-\tbl{Tuning Set and Testing Set}{
-\begin{tabular}{|l|l|c|c|}\hline
-Label & \multicolumn{1}{c|}{Description}
-& Number of Users &
-Number of Queries\\\hline
-Train70 & Training Data &
-\smash{\raise-7pt\hbox{70}} & 104\\
-\cline{1-2}\cline{4-4}
-Test70 & Testing Data I & & 105\\\hline
-Test30 & Testing Data II & 30 & 119\\\hline
-& Total & 100 & 328\\\hline
-\end{tabular}}
-\end{table}\end{verbatim}
-
-\begin{table}
-\tbl{Tuning Set and Testing Set}{
-\begin{tabular}{|l|l|c|c|}\hline
-Label & \multicolumn{1}{c|}{Description}
-& Number of Users &
-Number of Queries\\\hline
-Test 1 & Training Data &
-\smash{\raise-7pt\hbox{70}} & 104\\
-\cline{1-2}\cline{4-4}
-Test 2 & Testing Data I & & 105\\\hline
-Test 3 & Testing Data II & 30 & 119\\\hline
-& Total & 100 & 328\\\hline
-\end{tabular}}
-\end{table}
-\subsection{Landscaping Pages}
-\label{subsub:landscaping_pages}
-If a table is too wide to fit the standard measure, it may be turned,
-with its caption, to 90 degrees. Landscape tables cannot be produced
-directly using the \verb juliacon   class file because \TeX{} itself cannot
-turn the page, and not all device drivers provide such a facility.
-The following procedure can be used to produce such pages.
-\vskip 6pt
-Use the package \verb rotating   in your document and change the coding
-from
-\begin{verbatim}
-\begin{table}...\end{table}
-to
-\begin{sidewaystable}...\end{sidewaystable}
-and for figures
-\begin{figure}...\end{figure}
-to
-\begin{sidewaysfigure}...\end{sidewaysfigure}
-\end{verbatim}
-
-environments in your document to turn your table on the appropriate
-page of your document. For instance, the following code prints
-a page with the running head, a message half way down and the
-table number towards the bottom.
-\begin{verbatim}
-\begin{sidewaystable}
-\tbl{Landscape table caption to go here.}{...}
-\label{landtab}
-\end{sidewaystable}
-\end{verbatim}
-
-\subsection{Double Column Figure and Tables}
-\label{subsub:double_fig_tab}
-For generating the output of figures and tables in double column
-we can use the following coding:
-
-\begin{enumerate}
-\item For Figures:
-\begin{verbatim}
-\begin{figure*}...\end{figure*}
-\end{verbatim}
-\item For landscape figures:
-\begin{verbatim}
-\begin{sidewaysfigure*}...\end{sidewaysfigure*}
-\end{verbatim}
-\item For Tables:
-\begin{verbatim}
-\begin{table*}...\end{table*}
-\end{verbatim}
-\item For landscape tables:
-\begin{verbatim}
-\begin{sidewaystable*}...\end{sidewaystable*}
-\end{verbatim}
-\end{enumerate}
-
-\subsection{Typesetting Mathematics}
-\label{subsub:type_math}
-The \verb juliacon class file will set displayed mathematics with center to
-the column width, provided that you use the \LaTeXe{} standard of
-open and closed square brackets as delimiters.
-The equation
-\[
-\sum_{i=1}^p \lambda_i = (S)
-\]
-
-was typeset using the acmtog class file with the commands
-
-\begin{verbatim}
-\[
-\sum_{i=1}^p \lambda_i = (S)
-\]
-\end{verbatim}
-
-For display equations, cross-referencing is encouraged. For example,
-\begin{verbatim}
-\begin{equation}
-(n-1)^{-1} \sum^n_{i=1} (X_i - \overline{X})^2.
-\label{eq:samplevar}
-\end{equation}
-Equation~(\ref{eq:samplevar}) gives the formula for
-sample variance.
-\end{verbatim}
-The following output is generated with the above coding:
-\begin{equation}
-(n-1)^{-1} \sum^n_{i=1} (X_i - \overline{X})^2.
-\label{eq:samplevar}
-\end{equation}
-Equation~(\ref{eq:samplevar}) gives the formula for
-sample variance.
-
-\subsection{Enunciations}
-\label{subsub:enunciation}
-The \verb juliacon   class file generates the enunciations with the help of
-the following commands:
-\begin{verbatim}
-\begin{theorem}...\end{theorem}
-\begin{strategy}...\end{strategy}
-\begin{property}...\end{property}
-\begin{proposition}...\end{proposition}
-\begin{lemma}...\end{lemma}
-\begin{example}...\end{example}
-\begin{proof}...\end{proof}
-\begin{definition}...\end{definition}
-\begin{algorithm}...\end{algorithm}
-\begin{remark}...\end{remark}
-\end{verbatim}
-The above-mentioned coding can also include optional arguments
-such as
-\begin{verbatim}
-\begin{theorem}[...]. Example for theorem:
-\begin{theorem}[Generalized Poincare Conjecture]
-Four score and seven ... created equal.
-\end{theorem}
-\end{verbatim}
-
-\begin{theorem}[Generalized Poincare Conjecture]
-Four score and seven years ago our fathers brought forth,
-upon this continent, a new nation, conceived in Liberty,
- and dedicated to the proposition that all men are
-created equal.
-\end{theorem}
-
-\subsection{Extract}
-\label{subsub:extract}
-Extract environment should be coded within
-\begin{verbatim}
-\begin{extract}..\end{extract}
-\end{verbatim}
-
-\subsection{Balancing column at last page}
-\label{subsub:Balance}
-For balancing the both column length at last page use :
-\begin{verbatim}
-\vadjust{\vfill\pagebreak}
-\end{verbatim}
-
-%\vadjust{\vfill\pagebreak}
-
-at appropriate place in your \TeX{} file or in bibliography file.
-
-\section{Handling references}
-\label{subsub:references}
-References are most easily (and correctly) generated using the
-BIBTEX, which is easily invoked via
-\begin{verbatim}
-\bibliographystyle{juliacon}
-\bibliography{ref}
-\end{verbatim}
-When submitting the document source (.tex) file to external
-parties, the ref.bib file should be sent with it.
-\cite{bezanson2017julia}
-
 \input{bib.tex}

### 2019-05-30 paper/paper.tex
experiment plots
complexity curves

@@
 %
+\begin{figure}[htb]
+\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-8k.pdf}}
+\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-256k.pdf}}
+\caption{Empirical CDF from {\tt chipsort\_medium!} with 8k {\tt UInt32} values and {\tt chipsort\_large} with 256k values. Results from 10,000 measurements. Speedups are 14\% and 3\%.}
+\label{fig:bench-cdf}
+\end{figure}
+
+\begin{figure*}[htb]
+\centerline{\includegraphics[width=0.75\linewidth]{fig/chipsort-bench-curves.pdf}}
+\caption{Curves displaying the relative running time from each studied method. The ordinate represents the median of the measured running times divided by the input size at each test.}
+\label{fig:bench-curves}
+\end{figure*}
+
 \section{Conclusion}

### 2019-05-31 paper/paper.tex
final results, and starting to write experimens

@@
 
+Despite the interest due to parallelization, as the experiments show Combsort can display a great performance even in a serial version.
+
 \subsection{In-place matrix transpose}
@@
 %
-\begin{figure}[htb]
-\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-8k.pdf}}
-\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-256k.pdf}}
-\caption{Empirical CDF from {\tt chipsort\_medium!} with 8k {\tt UInt32} values and {\tt chipsort\_large} with 256k values. Results from 10,000 measurements. Speedups are 14\% and 3\%.}
-\label{fig:bench-cdf}
-\end{figure}
+
+\subsection{Small arrays}
+In our first experiment we evaluate the sorting of small arrays, with only 64 32-bit values.
+The ChipSort method based on sorting and bitonic merge networks was compared to Insertion sort and the mixed Comb and insertion sort.
+The small size of the array, and the small running time of the functions make it challenging to perform proper measurements.
+Our experimental task was therefore to sort 128 sequences of 64 contiguous values in memory, forming a single block of $2^13$ random entries.
+The in-place sorting functions were applied in sequence, in a for-loop.
+The results are in Table~\ref{tab:bench-small}.
+
+The {\tt chipsort\_small} function achieved a significant speed-up of over 3 times compared to Insertion sort or the Julia standard function, and twice over Combsort with Insertion sort.
+
+\begin{table}[h]
+\tabcolsep22pt
+\tbl{Sorting $128 \times 64$ {\tt UInt32} elements.}{
+\begin{tabular}{|l|r|}\hline
+Method & Median time\\\hline
+ChipSort small & 57.206 $\mu$s\\\hline
+Comb+Insertion & 133.510 $\mu$s\\\hline
+Insertion sort & 194.679 $\mu$s\\\hline
+Julia standard & 222.910 $\mu$s\\\hline
+\end{tabular}}
+\label{tab:bench-small}
+\end{table}
+
+The mixed Combsort plus Insertion sort works by proceeding like the regular Combsort until the interval size is unit, and then it switches to insertion sort.
+This is in a way similar to the Julia standard library, which is Quicksort switching to Insertion sort if the array is too small, with the difference that it is not a divide-and-conquer approach, and therefore in the mixed Combsort the final application of Insertion goes only once, and throught the whole array.
+This first experiment already demonstrates the potential of this method.
+Even though the parallelized function was clearly faster, the mixed Combsort also presented a speedup of at least 46\% compared to the other alternatives.
+
+\subsection{Medium and large arrays}
+In our second experiment we compared many different methods on inputs of exponentially increasing sizes, from $2^6$ up to $2^{20}$.
 
@@
 \end{figure*}
+
+\begin{figure}[htb]
+\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-8k.pdf}}
+\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-256k.pdf}}
+\caption{Empirical CDF from {\tt chipsort\_medium!} with 8k {\tt UInt32} values and {\tt chipsort\_large} with 256k values. Results from 10,000 measurements. Speedups are 7.3\% and 4.4\%.}
+\label{fig:bench-cdf}
+\end{figure}
+
+\begin{figure}[htb]
+\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-time.pdf}}
+\caption{Running time of the different sorting methods studied.}
+\label{fig:bench-cdf}
+\end{figure}
 

### 2019-05-31 paper/paper.tex
most of the results and nomenclature of ChipSort our favorite algorithm

--- 2019-05-30 paper/paper.tex
@@
 One of the most peculiar aspects of ChipSort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm.
-Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes insertion sort~\cite{dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}.
-The array is sweeped multiple times, and pairs of values at a given distance from each other are compared and stored in order.
+Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}.
+The array is swept multiple times, and pairs of values at a given distance from each other are compared and stored in order.
 The distance is reduced at each pass by a factor of $\frac{4}{3}$.
-This algorithm lends itself easily to vectorization: performing the comparisons over vectors of size $V$ effectively allows us to run the algorithm over $V$ lists simultaneouly.
+This algorithm lends itself easily to vectorization: performing the comparisons over vectors of size $V$ effectively allows us to run the algorithm over $V$ lists simultaneously.
 
-In practice, ChipSort performs a number of other operations before and after using Combsort. The {\tt chipsort\_medium!} function consists of the following steps:
+ChipSort contains two functions that utilize Combsort.
+The first is {\em chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the compiler optimizations.
+{\tt chipsort!} starts like Combsort until the interval size is 1, when it switches to Insertion sort.
+This is similar to what is often done in practical implementations of Quicksort, as in the Julia standard library, switching to Insertion sort when the array becomes too small.
+The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort is applied only once through the whole array.
 
+The other function is {\tt chipsort\_medium!}, an explicitly vectorized implementation utilizing {\tt SIMD.jl}.
+This second function also performs a number of other operations before and after using Combsort.
+It consists of the following steps:
 \begin{enumerate}
@@
 Each row is processed independently at first, and the transpose allows elements from distinct groups to interact.
+The last step is necessary to ensure the whole array is sorted, not viewed anymore as a matrix with independent rows.
 
-The last step is necessary to ensure the whole array is sorted, not vieved anymore as a matrix with independent rows.
-Insertion sort also presents a better performance than sticking to Combsort until the end, what would essentially mean using bubble sort once the interval becomes 1.
-Because the array is now approximately sorted, it is expected at this time that insertion sort does not have to move any out-of-order elements very far from where they are found.
+In both implementations switching to Insertion sort presents a better performance than sticking to Combsort until the end.
+It essentially means using Bubble sort once the interval becomes 1, therefore looking for an alternative seems like a good idea.
+Insertion sort should fare well because the array is now approximately sorted, and it should not have to move any out-of-order elements very far from where they are found.
+This utilization of Insertion sort as a final stage to Combsort does not seem to have been considered in the literature before, and may be the main contribution of the present work.
 
@@
 %
+This section reports experiments with the methods offered by the ChipSort library, as well as alternative baseline methods.
+The term {\it method} can be understood in this article as both the sorting algorithm and their actual implementations in the Julia language.
+The specific methods tested were:
+\begin{itemize}
+\item {\tt chipsort\_small!} in-register, based on sorting networks.
+\item {\tt chipsort\_medium!} vectorized Combsort, plus additional steps, finishing with Insertion sort.
+\item {\tt chipsort\_large} merge-sort with a multi-way merge tree, starting from chunks processed by {\tt chipsort\_medium!}.
+\item {\tt chipsort\_serial!} non-vectorized, serial Combsort finishing with Insertion sort when the interval becomes 1.
+\item {\tt insertion\_sort!} Insertion sort taken from the Julia library.
+\item {\tt sort!} the Julia standard sort, Quicksort finishing with Insertion sort for small arrays (20 elements).
+\end{itemize}
+
+The function names with a {\tt !} imply in-place operation.
+The output array allocation time for {\tt chipsort\_large} was not discounted in the experiments.
+While it may it not be irrelevant, it is small enough to be ignored with no effect to our conclusions.
+
+\subsection{Small arrays}
+In our first experiment we evaluate the sorting of small arrays, with only 64 32-bit values.
+The {\tt chipsort\_small} method based on sorting and bitonic merge networks was compared to serial Insertion sort, the Julia sort and {\tt chipsort!}.
+The small running time makes it challenging to perform proper measurements in this experiment.
+Our task was therefore to sort 128 sequences of 64 contiguous values in memory, forming a single block of $2^{13}$ random entries.
+The in-place sorting functions were applied in sequence, in a for-loop.
+The results are in Table~\ref{tab:bench-small}.
+
+The {\tt chipsort\_small} function achieved a significant speed-up of over 3 times compared to Insertion sort or the Julia standard function, and twice over {\tt chipsort!}.
+
+\begin{table}[h]
+\tabcolsep22pt
+\tbl{Sorting $128 \times 64$ {\tt UInt32} elements.}{
+\begin{tabular}{|l|r|}\hline
+Method & Median time\\\hline
+ChipSort small & 57.206 $\mu$s\\\hline
+ChipSort & 133.510 $\mu$s\\\hline
+Insertion sort & 194.679 $\mu$s\\\hline
+Julia standard & 222.910 $\mu$s\\\hline
+\end{tabular}}
+\label{tab:bench-small}
+\end{table}
+
+This first experiment already demonstrates the potential of {\tt chipsort!}, which presented a speedup of at least 46\% compared to the other alternatives, even though the function based on sorting networks was clearly faster.
+
+\subsection{Medium and large arrays}
+In our second experiment we compared many different methods on inputs of exponentially increasing sizes, from $2^6$ up to $2^{20}$.
+Figure~\ref{fig:bench-time} displays the resulting measurements.
+Although the curves look too concentrated in this visualization, it is possible to notice that {\tt chipsort!} consistently outperforms the Julia standard library {\tt sort!} by a factor of 80\% to 100\%, or up to twice as fast.
+We can also observe that Insertion sort soon becomes too inefficient.
+
 \begin{figure}[htb]
-\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-8k.pdf}}
-\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-256k.pdf}}
-\caption{Empirical CDF from {\tt chipsort\_medium!} with 8k {\tt UInt32} values and {\tt chipsort\_large} with 256k values. Results from 10,000 measurements. Speedups are 14\% and 3\%.}
-\label{fig:bench-cdf}
+\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-time.pdf}}
+\caption{Running time of the different sorting methods studied.}
+\label{fig:bench-time}
 \end{figure}
+
+Figure~\ref{fig:bench-curves} shows a different visualization from the same data in Figure~\ref{fig:bench-time}, with the times divided by the input size at each point.
+Here we can see more clearly how {\tt chipsort\_medium} and {\tt chipsort\_large} slightly outperform {\tt chipsort!} in a few cases.
 
@@
 \end{figure*}
+
+Figure~\ref{fig:bench-cdf} displays the measured times for
+
+ Speedups are 7.3\% and 4.4\%.
+
+\begin{figure}[htb]
+\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-8k.pdf}}
+\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-256k.pdf}}
+\caption{Empirical CDF from different methods with 8k and 256k {\tt UInt32} values. Results from 10,000 measurements.}
+\label{fig:bench-cdf}
+\end{figure}
 

### 2019-05-31 paper/paper.tex
more refs

@@
 One of the most peculiar aspects of ChipSort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm.
-Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}.
+Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376,INCERPI198737}.
 The array is swept multiple times, and pairs of values at a given distance from each other are compared and stored in order.
@@
 The first is {\em chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the compiler optimizations.
-{\tt chipsort!} starts like Combsort until the interval size is 1, when it switches to Insertion sort.
+{\tt chipsort!} starts with Combsort until the interval size is 1, when it switches to Insertion sort.
 This is similar to what is often done in practical implementations of Quicksort, as in the Julia standard library, switching to Insertion sort when the array becomes too small.
 The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort is applied only once through the whole array.
+Shell sort also terminates naturally applying Insertion sort to a partially sorted array.
 
@@
 Insertion sort should fare well because the array is now approximately sorted, and it should not have to move any out-of-order elements very far from where they are found.
-This utilization of Insertion sort as a final stage to Combsort does not seem to have been considered in the literature before, and may be the main contribution of the present work.
+
+Finishing a sorting method with Insertion sort is a common practice, present at least in Musser's Introsort~\cite{musser1997introspective}.
+The utilization of Insertion sort as a final stage to Combsort has been considered before in the literature~\cite{combwiki,INCERPI198737}, although this method does not seem to enjoy wide adoption.
+The main contribution of the present work may be to propose again this method in the SIMD context, extending AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, bringing experimental evidence of its good performance.
 
@@
 
-Figure~\ref{fig:bench-cdf} displays the measured times for
-
- Speedups are 7.3\% and 4.4\%.
+Figure~\ref{fig:bench-cdf} displays the measured times for the test cases with $2^{13}$ and $2^{18}$ elements.
+It is possible to see how all the ChipSort methods surpass the Julia standard library.
+In the $2^{13}$ case {\tt chipsort\_medium!} exhibits a slight speedup of 7.3\% over {\tt chipsort!}, and on the $2^{13}$ case it is {\tt chipsort\_large!} with 4.4\%.
 

### 2019-05-31 paper/paper.tex
starting the conclu

@@
 This section reports experiments with the methods offered by the ChipSort library, as well as alternative baseline methods.
-The term {\it method} can be understood in this article as both the sorting algorithm and their actual implementations in the Julia language.
-The specific methods tested were:
+The term {\it method} can be understood as both the sorting algorithm and their actual implementations in the Julia language.
+The specific methods tested are:
 \begin{itemize}
+\item {\tt chipsort!} Combsort finishing with Insertion sort, written as a serial program and vectorized by the compiler.
 \item {\tt chipsort\_small!} in-register, based on sorting networks.
@@
 \item {\tt chipsort\_large} merge-sort with a multi-way merge tree, starting from chunks processed by {\tt chipsort\_medium!}.
-\item {\tt chipsort\_serial!} non-vectorized, serial Combsort finishing with Insertion sort when the interval becomes 1.
 \item {\tt insertion\_sort!} Insertion sort taken from the Julia library.
@@
 
-The function names with a {\tt !} imply in-place operation.
-The output array allocation time for {\tt chipsort\_large} was not discounted in the experiments.
-While it may it not be irrelevant, it is small enough to be ignored with no effect to our conclusions.
+The {\tt !} implies in-place operation. The allocation time for {\tt chipsort\_large} was not discounted in our analyses.
 
@@
 The small running time makes it challenging to perform proper measurements in this experiment.
-Our task was therefore to sort 128 sequences of 64 contiguous values in memory, forming a single block of $2^{13}$ random entries.
+Our task was therefore to sort 128 sequences of 64 contiguous values in memory, forming a single block of $2^{13}$ uniform random entries.
 The in-place sorting functions were applied in sequence, in a for-loop.
 The results are in Table~\ref{tab:bench-small}.
-
-The {\tt chipsort\_small} function achieved a significant speed-up of over 3 times compared to Insertion sort or the Julia standard function, and twice over {\tt chipsort!}.
 
@@
 
-This first experiment already demonstrates the potential of {\tt chipsort!}, which presented a speedup of at least 46\% compared to the other alternatives, even though the function based on sorting networks was clearly faster.
+The {\tt chipsort\_small} function achieved a significant speed-up of over 3 times compared to Insertion sort or the Julia standard function, and twice over {\tt chipsort!}.
+Even though this in-register function was clearly faster, this first experiment already demonstrates the potential of {\tt chipsort!}, which presented a speedup of at least 46\% compared to the other alternatives.
 
@@
 
-Figure~\ref{fig:bench-cdf} displays the measured times for the test cases with $2^{13}$ and $2^{18}$ elements.
-It is possible to see how all the ChipSort methods surpass the Julia standard library.
+Figure~\ref{fig:bench-cdf} displays the measured times for $2^{13}$ and $2^{18}$ elements.
+All the tested ChipSort methods surpass the Julia standard library in these cases.
 In the $2^{13}$ case {\tt chipsort\_medium!} exhibits a slight speedup of 7.3\% over {\tt chipsort!}, and on the $2^{13}$ case it is {\tt chipsort\_large!} with 4.4\%.
@@
 %
+We have presented ChipSort, a Julia module for SIMD and cache-aware sorting.
+The main sorting method utilized in ChipSort is Combsort finalized with Insertion sort.
+This method is based on previous proposals for SIMD sorting~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} and other practical sorting methods~\cite{INCERPI198737,musser1997introspective}.
+This method can be implemented as a serial program that can be vectorized by the Julia compiler and LLVM, resulting in a sorting function up to twice as fast as the Julia standard library.
+This performance holds for at least a million 32-bit elements in a conventional computer.
+
+Other techniques available in ChipSort are sorting networks, bitonic merge networks, in-place matrix transpose and a multi-way merge tree.
+Vectorized in-register sorting, although limited to small powers of 2, proved to deliver speedups of up to 3 times relative to the standard library.
+For large arrays there have been limited results, only narrowly surpassing the main ChipSort method in few occasions.
+Merge-based techniques are expected to be more advantageous when sorting large registers, minimizing memory access.
+
+The main ChipSort method relies heavily on compiler optimizations, while other methods utilize explicit vectorization trough SIMD.jl.
+ChipSort makes extensive use of metaprogramming, in epecial Julia generated functions.
+This allows functions like sorting networks to be implemented in an abstract and generic way, based on representations of the networks as data structures used to assemble the final code by a high-level program.
+This contrasts with similar implementations in other libraries where networks are implemented straight as final functions.
+Julia also has the opportunity of performing many other optimizations since the library is implemented in Julia with abstract methods that can be specialized.
+
+The best result obtained in ChipSort is a simple serial function that gets vectorized.
 

### 2019-06-01 paper/paper.tex
first complete version, starting to review

@@
 \documentclass{juliacon}
+\usepackage{xspace}
 \setcounter{page}{1}
@@
 \begin{document}
+
+\newcommand{\chipsort}{{\tt ChipSort}\xspace}
 
@@
 
-Sorting is a fundamental programming problem with many important applications, being very useful to organize data for fast retrieval.
+Sorting is a fundamental programming problem with many important applications.
 While there are well-known sorting algorithms for conventional computers, especial techniques are required to take the most out of real hardware offering features such as cache memory and instruction-level parallelism.
@@
 It implements sorting networks and bitonic merge networks with SIMD instructions, with configurable vector sizes.
-It also implements Combsort, which lends itself easily to vectorization and achieves good performance in arrays that fit cache memory, with low penalty for random access.
-This also depends on a parallel in-place matrix transposition implemented in ChipSort.
+It also implements Combsort, which lends itself easily to vectorization and can achieve good performance depending on the memory access cost.
+Insertion sort is used to finalize.
 Large arrays are approached with a multi-way Mergesort.
 The implementation of ChipSort itself is of interest from a programming languages perspective due to its use of metaprogramming, and more specifically Julia's generated functions.
-This enables tuning the code generation to specific tasks and hardware at runtime, a feat owed to Julia.
-This article presents the implemented techniques as well as experiments that demonstrate the increased speed and the hardware adaptation capability of the module.
+This permits tuning the code to specific tasks and hardware at runtime, a feat owed to Julia.
+This article presents the implemented techniques as well as experiments that demonstrate speed gains compared to multiple standard libraries.
 
@@
 Sorting has enjoyed an important place within computer programming topics for a long time.
-Apart from having great importance for practical applications, it also attracts much attention as a theoretically interesting problem in itself~\cite{DBLP:books/lib/Knuth98a}.
+Apart from having great importance in practical applications, being very useful to organize data for fast retrieval, it also attracts attention as a theoretically interesting problem in itself~\cite{DBLP:books/lib/Knuth98a}.
 While sorting is a well-understood problem in general, with a few classic algorithms available, achieving optimal performance for specific problems and architectures may require a careful algorithm selection rather than just tuning parameters \cite[Part II, Introduction]{DBLP:books/daglib/0023376}.
@@
 The running time of a computer program is governed by many factors, starting with algorithmic complexity and the base clock speed of the processor.
-The observed performance in real computers has been increasingly depending in processor features that are not taken into account by the simplest models, though.
-These features include cache memory and paralellism at the processor and instruction level.
+The performance observed in real computers has been increasingly depending in processor features that are not taken into account by the simplest models, though.
+These features include cache memory and paralellism at the processor and instruction levels.
 Cache memory has become popular in consumer processors in the past couple of decades, allowing programs to often attain higher performance than would be possible with a simpler memory architecture, as long as memory access patterns exhibit temporal and spatial locality~\cite[Chapter 3]{Drepper07whatevery}.
-Writing software that can better cope with paralellism has also become necessary to exploit the full potential offered by modern and future processors~\cite{wilson2018}.
+Writing software that can better cope with parallelism has also become necessary to exploit the full potential offered by modern and future processors~\cite{wilson2018}.
 
-The interest in sorting and the great relevance of cache memory and paralellism in modern computing naturally lead to research on sorting algorithms that account for these factors.
-Processors offering SIMD (single instruction multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
+The interest in sorting and the great relevance of cache memory and parallelism in modern computing naturally led to research on sorting algorithms that account for these factors.
+Processors offering SIMD (single instruction, multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
 One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}.
-The current availability of processors with 512 bit registers underscore the need of being able to write software that exploit vector operations, not to mention GPUs.
+The current availability of processors with 512 bit registers underscores the need of being able to write software that exploits vector operations, not to mention GPUs.
 
-At the time CPUs with SIMD instructions became widely available there was already a trend to employ specialized hardware like GPUs for scientific applications~\cite{larsen2001fast,DBLP:conf/micro/ThompsonHO02}.
-Althoug SIMD was used mostly for multimedia applications at first~\cite{CHEN2006509,DBLP:journals/mam/SlingerlandS05}, we can cite some early alternative uses including 3D graphics~\cite{DBLP:conf/pcm/MaY02} and machine learning~\cite{DBLP:conf/europar/StreyB01}.
+At the time SIMD CPUs became widely available there was already a trend to employ specialized hardware like GPUs for scientific applications~\cite{larsen2001fast,DBLP:conf/micro/ThompsonHO02}.
+Although SIMD was used mostly for multimedia applications at first~\cite{CHEN2006509,DBLP:journals/mam/SlingerlandS05}, some early alternative uses include 3D graphics~\cite{DBLP:conf/pcm/MaY02} and machine learning~\cite{DBLP:conf/europar/StreyB01}.
 
-Apart from earlier work on parallel sorting in other architectures, the late 2000s saw the first practical demonstrations of sorting with modern consumer SIMD chips~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
+Following earlier work on parallel sorting in other architectures, the late 2000s saw the first practical demonstrations of sorting with modern consumer SIMD chips~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
 Most of the techniques utilized then are still employed in more recent works~\cite{DBLP:journals/pvldb/BalkesenATO13,DBLP:journals/pvldb/InoueT15}.
 
-This article presents \verb ChipSort, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
+This article presents \chipsort, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
 These include sorting and merging networks, the use of Combsort for medium sized arrays and of multi-way merging for large arrays.
 These techniques require not only tuning parameters such as vector and array sizes, but also generating specific code when implementing different size networks.
-\verb ChipSort seeks to generate efficient custom code for each task, relying on Julia metaprogramming capabilities for that.
+\chipsort seeks to generate efficient custom code for each task, relying on Julia metaprogramming capabilities for that.
 
-Following Section~\ref{sec:methods} presents the techniques employed in ChipSort.
+Section~\ref{sec:methods} ahead presents the techniques employed in \chipsort.
 Section~\ref{sec:experiments} presents experimental results assessing the performance of the module and Section~\ref{sec:conclusion} brings some concluding remarks.
@@
 %
-ChipSort offers three high-level functions that can be used in applications as replacements for other sorting functions such as \verb sort  in the Julia standard library.
-Additional work is required from the user selecting among the three functions and also setting some numeric parameters.
-The functions are intended to be used with arrays of different sizes, and \verb chipsort_large  offers a good overview of all the techniques implemented in the module.
+\chipsort offers four high-level functions that can be used in applications as replacements for other sorting functions such as \verb sort  in the Julia standard library.
+Additional work may be required from the user selecting among the functions and also setting some numeric parameters.
+The {\tt chipsort!} function does not require any configuration and offers a great performance in a wide range of cases.
+The remaining functions are intended to be used with arrays of different sizes, and \verb chipsort_large  offers a good overview of all the techniques implemented in the module.
 Some of these techniques might also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.
-ChipSort does not currently support registers with separate key and payload, what might allow significant optimizations.
-Only the sorting of simple data types such as 32 and 64-bit integers and floating-point numbers were studied.
+\chipsort does not currently support data records with separate key and payload, what might allow significant optimizations.
+Only the sorting of simple data types such as 32 and 64-bit integers and floating-point numbers has been studied so far.
 
@@
 
-In ChipSort, as in the related work, parallelism is mainly exploited by carrying out the comparisons of the sorting network over vectors of size $V$ contained in SIMD registers, effectively running the network $V$ times simultaneously.
+In \chipsort, as in the related work, parallelism is mainly exploited by carrying out the comparisons of the sorting network over vectors of size $V$ contained in SIMD registers, effectively running the network $V$ times simultaneously.
 Other forms of instruction-level parallelism can also be in play, though.
@@
 
-ChipSort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/ChipSort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}.
-The generated function \href{https://github.com/nlw0/ChipSort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} works by assigning each element of the sequence at each step to a variable, and calculating values from the next step according to the network specification.
+\chipsort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}.
+The generated function \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} works by assigning each element of the sequence at each step to a variable, and calculating values from the next step according to the network specification.
 The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types.
@@
 
-This function is a good first example of the utilization of Julia's meta-programming features in ChipSort.
+This function is a good first example of the utilization of Julia's meta-programming features in \chipsort.
 Multiple implementations of sorting networks are available~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08,sortingnetworksjl,ultrasort}, however they implement networks of different sizes straight as code.
-In ChipSort they are represented as data structures that guide a generated function that produces equivalent code.
+In \chipsort they are represented as data structures that guide a generated function that produces equivalent code.
 While this data structure currently is hard-coded, it can potentially be produced only when necessary by an implementation of an algorithm such as Bose-Nelson, what remains a future plan for the module.
@@
 It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works.
-In practice ChipSort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}.
+In practice \chipsort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}.
 This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers.
@@
 This data must be transposed to produce sorted vectors.
-This is done in ChipSort through {\tt transpose\_vecs}, which is a generated function that supports rectangular matrices, but with dimensions constrained to powers of 2.
+This is done in \chipsort through {\tt transpose\_vecs}, which is a generated function that supports rectangular matrices, but with dimensions constrained to powers of 2.
 The generated code consists mostly of calls to {\tt SIMD.shufflevec}.
-This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions, contemplating each different case, and in ChipSort there is only a single higher-level Julia generated function that produces equivalent code.
+This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions, contemplating each different case, and in \chipsort there is only a single higher-level Julia generated function that produces equivalent code.
 
@@
 A bitonic sequence starts either increasing or decreasing and contains at most one change of direction.
-Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in ChipSort in a single generated function, {\tt bitonic\_merge}.
+Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, {\tt bitonic\_merge}.
 Merging two sorted vectors requires first reversing one of them, and the result is stored in two new vectors of the same size containing the first and second halves of the total sorted sequence.
@@
 Apart from the use in sorting small sequences and in the multi-way merge tree, bitonic merge networks can also be used to implement a regular merge sort.
-This technique was investigated in ChipSort only superficially, and while there were no promising results in terms of performance, ChipSort also contains the generated function {\tt bitonic\_merge\_interleaved} to illustrate how metaprogramming can help implementing multiple simultaneous merges with interleaved execution, as done in other projects~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
+This technique was investigated in \chipsort only superficially, and while there were no promising results in terms of performance, \chipsort also contains the generated function {\tt bitonic\_merge\_interleaved} to illustrate how metaprogramming can help implementing multiple simultaneous merges with interleaved execution, as done in other projects~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
 
@@
 %
-One of the most peculiar aspects of ChipSort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm.
+One of the most peculiar aspects of \chipsort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm.
 Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376,INCERPI198737}.
@@
 
-ChipSort contains two functions that utilize Combsort.
+\chipsort contains two functions that utilize Combsort.
 The first is {\em chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the compiler optimizations.
@@
 
-ChipSort contains the generated function {\tt transpose!} to perform in-place matrix transposition.
+\chipsort contains the generated function {\tt transpose!} to perform in-place matrix transposition.
 The cycle seeds are computed at the time of code generation, and at runtime the function only needs to move the data and compute the indices detecting when each cycle finishes.
@@
 
-The intermediate data from the tree is kept by ChipSort in a contiguous array, treated as a heap.
+The intermediate data from the tree is kept by \chipsort in a contiguous array, treated as a heap.
 Each node ancestor can be found simply by the Euclidean division of its index by 2.
@@
 %
-This section reports experiments with the methods offered by the ChipSort library, as well as alternative baseline methods.
+This section reports experiments with the methods offered by the \chipsort library, as well as alternative baseline methods.
 The term {\it method} can be understood as both the sorting algorithm and their actual implementations in the Julia language.
@@
 Method & Median time\\\hline
-ChipSort small & 57.206 $\mu$s\\\hline
-ChipSort & 133.510 $\mu$s\\\hline
+\chipsort small & 57.206 $\mu$s\\\hline
+\chipsort & 133.510 $\mu$s\\\hline
 Insertion sort & 194.679 $\mu$s\\\hline
@@
 The {\tt chipsort\_small} function achieved a significant speed-up of over 3 times compared to Insertion sort or the Julia standard function, and twice over {\tt chipsort!}.
-Even though this in-register function was clearly faster, this first experiment already demonstrates the potential of {\tt chipsort!}, which presented a speedup of at least 46\% compared to the other alternatives.
+Even though the in-register function was clearly faster, this first experiment already demonstrates the potential of {\tt chipsort!}, which presented a speedup of at least 46\% compared to the other alternatives.
 
@@
 Figure~\ref{fig:bench-cdf} displays the measured times for $2^{13}$ and $2^{18}$ elements.
-All the tested ChipSort methods surpass the Julia standard library in these cases.
+All the tested \chipsort methods surpass the Julia standard library in these cases.
 In the $2^{13}$ case {\tt chipsort\_medium!} exhibits a slight speedup of 7.3\% over {\tt chipsort!}, and on the $2^{13}$ case it is {\tt chipsort\_large!} with 4.4\%.
@@
 
+A simple experiment was also carried out with other programming languages in the same machine to provide a richer context.
+Table~\ref{tab:bench-cxx} diplays statistics collected from Julia, Python and C++ programs.
+For the case of Python the value obtained when calling the function through PyCall was better than measuring in the native Python shell, therefore the best value was kept.
+Results obtained for C++ with Julia interopt using Cxx.jl were slightly worse than native, thus discarded.
+While Numpy was almost as fast as Julia, no reason could be found for the significantly worse performance from the C++ standard libraries.
+Supposedly there are important optimizations missing.
+
+\begin{table}[h]
+\tabcolsep11pt
+\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different platforms.}{
+\begin{tabular}{|l|r|r|}\hline
+Method & 8k & 1M\\\hline
+\chipsort & 172.35 $\mu$s & 35.28 ms\\\hline
+Julia standard & 345.46 $\mu$s & 68.39 ms\\\hline
+Numpy via PyCall & 369.50 $\mu$s & 70.01 ms\\\hline
+C++ {\tt std::qsort} & 788.59 $\mu$s & 147.37 ms\\\hline
+C++ {\tt std::sort} & 1,207.23 $\mu$s & 224.71 ms\\\hline
+\end{tabular}}
+\label{tab:bench-cxx}
+\end{table}
+
 \section{Conclusion}
@@
 %
-We have presented ChipSort, a Julia module for SIMD and cache-aware sorting.
-The main sorting method utilized in ChipSort is Combsort finalized with Insertion sort.
+We have presented \chipsort, a Julia module for SIMD and cache-aware sorting.
+The main sorting method utilized in \chipsort is Combsort finalized with Insertion sort.
 This method is based on previous proposals for SIMD sorting~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} and other practical sorting methods~\cite{INCERPI198737,musser1997introspective}.
-This method can be implemented as a serial program that can be vectorized by the Julia compiler and LLVM, resulting in a sorting function up to twice as fast as the Julia standard library.
-This performance holds for at least a million 32-bit elements in a conventional computer.
+This method was implemented as a serial program vectorized by the Julia compiler and LLVM, resulting in a sorting function up to twice as fast as the Julia standard library sort.
+This performance holds for at least a million 32-bit integers in a conventional computer.
 
-Other techniques available in ChipSort are sorting networks, bitonic merge networks, in-place matrix transpose and a multi-way merge tree.
+Other techniques available in \chipsort are sorting networks, bitonic merge networks, in-place matrix transpose and a multi-way merge tree.
 Vectorized in-register sorting, although limited to small powers of 2, proved to deliver speedups of up to 3 times relative to the standard library.
-For large arrays there have been limited results, only narrowly surpassing the main ChipSort method in few occasions.
-Merge-based techniques are expected to be more advantageous when sorting large registers, minimizing memory access.
 
-The main ChipSort method relies heavily on compiler optimizations, while other methods utilize explicit vectorization trough SIMD.jl.
-ChipSort makes extensive use of metaprogramming, in epecial Julia generated functions.
-This allows functions like sorting networks to be implemented in an abstract and generic way, based on representations of the networks as data structures used to assemble the final code by a high-level program.
-This contrasts with similar implementations in other libraries where networks are implemented straight as final functions.
-Julia also has the opportunity of performing many other optimizations since the library is implemented in Julia with abstract methods that can be specialized.
+Results were limited for the method dedicated to large arrays, only narrowly surpassing the main \chipsort method in few occasions.
+Merge-based techniques may be more advantageous when sorting large registers, though.
 
-The best result obtained in ChipSort is a simple serial function that gets vectorized.
+The main \chipsort method relies heavily on compiler optimizations, while other methods utilize explicit vectorization trough SIMD.jl.
+\chipsort makes extensive use of metaprogramming, in especial Julia generated functions.
+This allows functions like sorting networks to be implemented in an abstract and generic way.
+The networks are represented as data structures, used by a higher-level program to assemble the final network code.
+This contrasts with other libraries where different networks are implemented straight as final functions.
+Julia also has the opportunity to performing many optimizations since \chipsort is implemented in pure Julia with abstract methods that can be specialized.
+
+It is a testament to the quality of the Julia compiler infrastructure that the most successful method implemented in \chipsort has no explicit vectorization or intrinsics.
+It is just a simple, scalar implementation of Combsort that is optimized by the compiler to generate code adapted to different problem settings and to the hardware architecture at run-time.
+In the cases where explicit vectorization can produce better code, though, Julia still offers a suitable framework that allowed \chipsort to implement sophisticated techniques with concise code.
+And while this project can illustrate how well Julia implements traditional language features such as parametric types, its metaprogramming features make it stand out from other languages even more.
+
+Some future prospects for \chipsort are to better deal with floating-point comparisons; implement stable sorting of key-register data; explore the limits of in-register processing, especially in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to produce optimized code for complex tasks such as multi-stage multi-threaded merge sorting of large arrays.
 

### 2019-06-01 paper/paper.tex
1st rev on the way...

@@
 \chipsort offers four high-level functions that can be used in applications as replacements for other sorting functions such as \verb sort  in the Julia standard library.
-Additional work may be required from the user selecting among the functions and also setting some numeric parameters.
-The {\tt chipsort!} function does not require any configuration and offers a great performance in a wide range of cases.
-The remaining functions are intended to be used with arrays of different sizes, and \verb chipsort_large  offers a good overview of all the techniques implemented in the module.
+The {\tt chipsort!} function does not require any configuration and offers a great performance on a wide range of cases.
+The remaining functions are intended to arrays of different sizes, and require choosing some numeric parameters.
+The function \verb chipsort_large  offers a good overview of all the techniques implemented in the module, and this section is structured according to its stages.
 Some of these techniques might also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.
-\chipsort does not currently support data records with separate key and payload, what might allow significant optimizations.
-Only the sorting of simple data types such as 32 and 64-bit integers and floating-point numbers has been studied so far.
 
-This section will present how \verb chipsort_large  works, detailing each step of the process.
-It mostly follows~\cite{DBLP:journals/pvldb/InoueT15} with a few modifications.
-This function achieves sorting by first splitting the input array in chunks that are sorted separately with \verb chipsort_medium!
+The way \chipsort  works mostly follows~\cite{DBLP:journals/pvldb/InoueT15}, although \chipsort does not support yet registers with key and payload.
+Another main difference is that Combsort is always employed with an Insertion sort finalization.
+
+The function \verb chipsort_large  achieves sorting by first splitting the input array in chunks that are sorted separately with \verb chipsort_medium!
 and then merged simultaneously.
@@
 
-The remainder of this section will detail the multiple techniques employed in this process.
+The remainder of this section details the multiple techniques employed in this process.
 
@@
 A sorting network can be represented by a Knuth graph such as Figure~\ref{fig:sorting-network}.
+Each vertical line is a comparator.
 The graph represents a network that takes an array of four values $(A_1, A_2, A_3, A_4)$ and outputs the sorted sequence $(S_1, S_2, S_3, S_4)$.
-Each vertical line in the graph represents a comparator.
 \begin{figure}[htb]
@@
 
-In \chipsort, as in the related work, parallelism is mainly exploited by carrying out the comparisons of the sorting network over vectors of size $V$ contained in SIMD registers, effectively running the network $V$ times simultaneously.
+In \chipsort, as in the related work, parallelism is mainly exploited by carrying out the comparisons of the sorting network over vectors of size $V$ contained in SIMD registers, effectively running $V$ networks simultaneously.
 Other forms of instruction-level parallelism can also be in play, though.
-In especial, it may also be possible to parallelize instructions such as \verb min  and \verb max , and also the loading of data into registers.
-This is often done implicitly by the microarchitecture, and programmers can only try to make sure their code is suitable for this by properly ordering operations and avoiding branching, for instance.
+In especial, it may be possible to parallelize instructions such as \verb min  and \verb max , or the loading of data into registers.
+This is often done implicitly by the microarchitecture, and programmers can only make sure their code is suitable for this by properly ordering operations and avoiding branching, for instance.
 Part of this work is expected to be performed by the compiler, and especially LLVM in the case of Julia.
@@
 The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types.
-Figure~\ref{fig:sort-net-listing}
+Figure~\ref{fig:sort-net-listing} shows one example of generated code.
 
@@
 
-This function is a good first example of the utilization of Julia's meta-programming features in \chipsort.
-Multiple implementations of sorting networks are available~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08,sortingnetworksjl,ultrasort}, however they implement networks of different sizes straight as code.
-In \chipsort they are represented as data structures that guide a generated function that produces equivalent code.
-While this data structure currently is hard-coded, it can potentially be produced only when necessary by an implementation of an algorithm such as Bose-Nelson, what remains a future plan for the module.
+This function is a good first example of how \chipsort utilizes Julia's metaprogramming features.
+Multiple implementations of sorting networks are available in the literature~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08,sortingnetworksjl,ultrasort}, however they implement networks of different sizes straight as code.
+In \chipsort they are represented as data structures that guide a generated function to produce equivalent code.
+While this data structure is currently hard-coded, it could be potentially produced only when necessary by an algorithm such as Bose-Nelson, what remains a future plan for the module.
 
@@
 In practice \chipsort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}.
-This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers.
+This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers, or spilling data to the stack if all registers are occupied.
 
@@
 The generated code consists mostly of calls to {\tt SIMD.shufflevec}.
-This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions, contemplating each different case, and in \chipsort there is only a single higher-level Julia generated function that produces equivalent code.
+This is another instance of an operation that other projects~\cite{DBLP:journals/pvldb/BalkesenATO13,ultrasort} implement with multiple low-level functions contemplating each different case, while in \chipsort there is only a single Julia generated function that produces equivalent code.
 
-Figure~\ref{fig:transpose-vecs} displays an example of 64 random integers loaded into SIMD registers as 4 vectors.
+Figure~\ref{fig:transpose-vecs} displays an example of 16 random integers loaded into SIMD registers as 4 vectors.
 After {\tt sort\_vecs} each lane contains a sorted sequence, and {\tt transpose\_vecs} turns the lanes into vectors (registers).
@@
 A bitonic sequence starts either increasing or decreasing and contains at most one change of direction.
-Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, {\tt bitonic\_merge}.
-Merging two sorted vectors requires first reversing one of them, and the result is stored in two new vectors of the same size containing the first and second halves of the total sorted sequence.
+Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/bitonic-merge-network.jl#L12}{\tt bitonic\_merge}.
+Merging two sorted vectors requires first reversing one of them, and the result is stored into two new vectors of the same size containing the first and second halves of the total sorted sequence.
 
@@
 One of the most peculiar aspects of \chipsort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm.
-Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376,INCERPI198737}.
+Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{INCERPI198737,dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}.
 The array is swept multiple times, and pairs of values at a given distance from each other are compared and stored in order.
 The distance is reduced at each pass by a factor of $\frac{4}{3}$.
-This algorithm lends itself easily to vectorization: performing the comparisons over vectors of size $V$ effectively allows us to run the algorithm over $V$ lists simultaneously.
+This algorithm lends itself easily to vectorization, either by parallelizing the operations from a single array or executing the algorithm over independent arrays in parallel.
 
-\chipsort contains two functions that utilize Combsort.
-The first is {\em chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the compiler optimizations.
+\chipsort contains two functions that utilize Combsort, employing parallelism in these two distinct forms.
+The first is {\tt chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the optimizing compiler.
 {\tt chipsort!} starts with Combsort until the interval size is 1, when it switches to Insertion sort.
@@
 The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort is applied only once through the whole array.
-Shell sort also terminates naturally applying Insertion sort to a partially sorted array.
 
 The other function is {\tt chipsort\_medium!}, an explicitly vectorized implementation utilizing {\tt SIMD.jl}.
-This second function also performs a number of other operations before and after using Combsort.
+This second function also performs a number of operations before and after Combsort.
 It consists of the following steps:
@@
 Finishing a sorting method with Insertion sort is a common practice, present at least in Musser's Introsort~\cite{musser1997introspective}.
-The utilization of Insertion sort as a final stage to Combsort has been considered before in the literature~\cite{combwiki,INCERPI198737}, although this method does not seem to enjoy wide adoption.
-The main contribution of the present work may be to propose again this method in the SIMD context, extending AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, bringing experimental evidence of its good performance.
+The utilization of Insertion sort as a final stage to Combsort has been considered before~\cite{combwiki,INCERPI198737}, although the method does not seem to enjoy wide adoption.
+The main contribution of the present work may be to propose it again in the SIMD context, extending AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} and bringing experimental evidence of its good performance.
 
@@
 In-place matrix transposition can be attained by moving a number to its destination, then moving the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}.
-While carrying out this procedure is simple, it is not trivial to find out which are the cycles to tranpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[Sec.
+While carrying out this procedure is simple, it is not trivial to find out what are the cycles to tranpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[Sec.
 1.3.3, Ex.
@@
 
-\chipsort contains the generated function {\tt transpose!} to perform in-place matrix transposition.
+\chipsort contains the generated function \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/transpose-vecs.jl#L19}{\tt transpose!} to perform in-place matrix transposition.
 The cycle seeds are computed at the time of code generation, and at runtime the function only needs to move the data and compute the indices detecting when each cycle finishes.
@@
 The intermediate data from the tree is kept by \chipsort in a contiguous array, treated as a heap.
-Each node ancestor can be found simply by the Euclidean division of its index by 2.
+Assuming 1-based indexing, each node ancestor can be found simply by the Euclidean division of its index by 2.
 During operation it is assumed this data should fit some level of cache memory, as in the related work~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15}.
@@
 This section reports experiments with the methods offered by the \chipsort library, as well as alternative baseline methods.
-The term {\it method} can be understood as both the sorting algorithm and their actual implementations in the Julia language.
+The term {\it method} can be understood as both the algorithms and their implementations in the Julia language.
 The specific methods tested are:
@@
 
+\subsection{Other languages}
 A simple experiment was also carried out with other programming languages in the same machine to provide a richer context.
@@
 For the case of Python the value obtained when calling the function through PyCall was better than measuring in the native Python shell, therefore the best value was kept.
-Results obtained for C++ with Julia interopt using Cxx.jl were slightly worse than native, thus discarded.
+Results obtained for C++ with Julia interoperation using Cxx.jl were slightly worse than native, thus discarded.
 While Numpy was almost as fast as Julia, no reason could be found for the significantly worse performance from the C++ standard libraries.

### 2019-06-01 paper/paper.tex
revisada geral 1

@@
 Sorting is a fundamental programming problem with many important applications.
-While there are well-known sorting algorithms for conventional computers, especial techniques are required to take the most out of real hardware offering features such as cache memory and instruction-level parallelism.
+While there are well-known sorting algorithms for conventional computers, special techniques are required to take the most out of real hardware offering features such as cache memory and instruction-level parallelism.
 ChipSort is a Julia module for SIMD and cache-aware sorting.
@@
 Other forms of instruction-level parallelism can also be in play, though.
-In especial, it may be possible to parallelize instructions such as \verb min  and \verb max , or the loading of data into registers.
+It may be possible to parallelize instructions such as \verb min  and \verb max , or the loading of data into registers.
 This is often done implicitly by the microarchitecture, and programmers can only make sure their code is suitable for this by properly ordering operations and avoiding branching, for instance.
-Part of this work is expected to be performed by the compiler, and especially LLVM in the case of Julia.
+Part of this work is expected to be performed by the compiler, and in particular LLVM in the case of Julia.
 
@@
 %
-This section reports experiments with the methods offered by the \chipsort library, as well as alternative baseline methods.
+This section reports experiments with the \chipsort methods, as well as alternative ones.
 The term {\it method} can be understood as both the algorithms and their implementations in the Julia language.
@@
 \subsection{Small arrays}
-In our first experiment we evaluate the sorting of small arrays, with only 64 32-bit values.
-The {\tt chipsort\_small} method based on sorting and bitonic merge networks was compared to serial Insertion sort, the Julia sort and {\tt chipsort!}.
+In our first experiment we evaluate the sorting of small arrays, with 64 32-bit values.
+The {\tt chipsort\_small!} method based on sorting and bitonic merge networks was compared to serial Insertion sort, the Julia sort and {\tt chipsort!}.
 The small running time makes it challenging to perform proper measurements in this experiment.
 Our task was therefore to sort 128 sequences of 64 contiguous values in memory, forming a single block of $2^{13}$ uniform random entries.
-The in-place sorting functions were applied in sequence, in a for-loop.
+Sorting was then performed in-place, in a for-loop.
 The results are in Table~\ref{tab:bench-small}.
@@
 
-The {\tt chipsort\_small} function achieved a significant speed-up of over 3 times compared to Insertion sort or the Julia standard function, and twice over {\tt chipsort!}.
+The {\tt chipsort\_small} function achieved a significant acceleration of over 3 times relative to Insertion sort or the Julia standard function, and twice over {\tt chipsort!}.
 Even though the in-register function was clearly faster, this first experiment already demonstrates the potential of {\tt chipsort!}, which presented a speedup of at least 46\% compared to the other alternatives.
@@
 Figure~\ref{fig:bench-curves} shows a different visualization from the same data in Figure~\ref{fig:bench-time}, with the times divided by the input size at each point.
-Here we can see more clearly how {\tt chipsort\_medium} and {\tt chipsort\_large} slightly outperform {\tt chipsort!} in a few cases.
+The vertical dashed lines indicate where the input array would be as big as each cache level.
+Here we can see more clearly how {\tt chipsort\_medium!} and {\tt chipsort\_large} slightly outperform {\tt chipsort!} in a few cases.
 
@@
 A simple experiment was also carried out with other programming languages in the same machine to provide a richer context.
-Table~\ref{tab:bench-cxx} diplays statistics collected from Julia, Python and C++ programs.
-For the case of Python the value obtained when calling the function through PyCall was better than measuring in the native Python shell, therefore the best value was kept.
-Results obtained for C++ with Julia interoperation using Cxx.jl were slightly worse than native, thus discarded.
-While Numpy was almost as fast as Julia, no reason could be found for the significantly worse performance from the C++ standard libraries.
-Supposedly there are important optimizations missing.
+Table~\ref{tab:bench-cxx} displays statistics collected from Julia, Python and C++ programs.
+For the case of Python the measurements obtained by calling the function through PyCall were better than using the native Python shell, therefore the best value was kept.
+Results obtained for C++ with Julia inter-operation using Cxx.jl were slightly worse than native, thus discarded.
+While Numpy was almost as fast as Julia, no reason could be found for the dismal performance from the C++ standard library.
 
@@
 
-Results were limited for the method dedicated to large arrays, only narrowly surpassing the main \chipsort method in few occasions.
-Merge-based techniques may be more advantageous when sorting large registers, though.
+Results were limited for the method dedicated to large arrays, only narrowly surpassing the main \chipsort method in few cases.
+Merge-based techniques may be more advantageous when sorting large records, though.
 
-The main \chipsort method relies heavily on compiler optimizations, while other methods utilize explicit vectorization trough SIMD.jl.
-\chipsort makes extensive use of metaprogramming, in especial Julia generated functions.
+While the main \chipsort method relies on compiler optimizations, other methods utilize explicit vectorization trough SIMD.jl.
+\chipsort makes extensive use of metaprogramming, more specifically Julia generated functions.
 This allows functions like sorting networks to be implemented in an abstract and generic way.
@@
 This contrasts with other libraries where different networks are implemented straight as final functions.
-Julia also has the opportunity to performing many optimizations since \chipsort is implemented in pure Julia with abstract methods that can be specialized.
+Julia also has the opportunity to perform many optimizations since \chipsort is implemented in pure Julia with abstract methods that can be specialized.
 
@@
 It is just a simple, scalar implementation of Combsort that is optimized by the compiler to generate code adapted to different problem settings and to the hardware architecture at run-time.
-In the cases where explicit vectorization can produce better code, though, Julia still offers a suitable framework that allowed \chipsort to implement sophisticated techniques with concise code.
-And while this project can illustrate how well Julia implements traditional language features such as parametric types, its metaprogramming features make it stand out from other languages even more.
+And in the cases where explicit vectorization can produce better code Julia still offers a suitable framework that allowed \chipsort to implement sophisticated techniques with concise code.
+This project can illustrate how well Julia implements traditional language features such as parametric types, and its metaprogramming features make it stand out from other languages even more.
 
-Some future prospects for \chipsort are to better deal with floating-point comparisons; implement stable sorting of key-register data; explore the limits of in-register processing, especially in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to produce optimized code for complex tasks such as multi-stage multi-threaded merge sorting of large arrays.
+Some future prospects for \chipsort are to better deal with floating-point comparisons; implement stable sorting of key-payload data; explore the limits of in-register processing, specially in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to produce optimized code for complex tasks such as multi-stage multi-threaded merge sorting of large arrays.
 

### 2019-06-02 paper/paper.tex
just one more experiment

@@
 Large arrays are approached with a multi-way Mergesort.
-The implementation of ChipSort itself is of interest from a programming languages perspective due to its use of metaprogramming, and more specifically Julia's generated functions.
-This permits tuning the code to specific tasks and hardware at runtime, a feat owed to Julia.
+The implementation of ChipSort itself is of interest from a programming languages perspective due to its use of metaprogramming, more specifically Julia's generated functions.
+This enables custom code generation for different tasks and hardware at run-time, a feat owed to Julia.
 This article presents the implemented techniques as well as experiments that demonstrate speed gains compared to multiple standard libraries.
@@
 \begin{figure}[htb]
-\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-time.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-time.pdf}}
 \caption{Running time of the different sorting methods studied.}
@@
 
-\begin{figure*}[htb]
-\centerline{\includegraphics[width=0.75\linewidth]{fig/chipsort-bench-curves.pdf}}
+\begin{figure*}[thb]
+\centerline{\includegraphics[width=0.7\linewidth]{fig/chipsort-bench-curves.pdf}}
 \caption{Curves displaying the relative running time from each studied method. The ordinate represents the median of the measured running times divided by the input size at each test.}
@@
 \begin{figure}[htb]
-\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-8k.pdf}}
-\centerline{\includegraphics[width=0.99\linewidth]{fig/chipsort-bench-256k.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-8k.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-256k.pdf}}
 \caption{Empirical CDF from different methods with 8k and 256k {\tt UInt32} values. Results from 10,000 measurements.}
@@
 \subsection{Other languages}
-A simple experiment was also carried out with other programming languages in the same machine to provide a richer context.
+A simple experiment was also carried out with different programming languages in the same machine, and with Julia in different machines to provide a richer context.
 Table~\ref{tab:bench-cxx} displays statistics collected from Julia, Python and C++ programs.
@@
 \end{table}
+
+Table~\ref{tab:bench-512} contains the results for {\tt chipsort!} and the Julia standard {\tt sort!} in an Intel processor with AVX512 instructions, and an ARM processor with NEON instructions.
+\begin{table}[h]
+\tabcolsep11pt
+\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different platforms.}{
+\begin{tabular}{|l|r|r|}\hline
+Method & 8k & 1M\\\hline
+AXV512 \chipsort & 177.4 $\mu$s & 35.20 ms\\\hline
+AXV512 Julia standard & 347.5 $\mu$s & 69.07 ms\\\hline
+NEON \chipsort & 350.9 $\mu$s & 73.36 ms\\\hline
+NEON Julia standard & 573.2 $\mu$s & 113.68 ms\\\hline
+\end{tabular}}
+\label{tab:bench-512}
+\end{table}
+
+No change in performance can be seen in the AVX512 machine, and on the ARM machine the speedup was smaller than before: only 60\%.
+The most notable fact in this experiment is the fact the same code was successfully employed to produce a vectorized sort on machines with different vector sizes and even instruction sets, always with beneficial effects.
 

### 2019-06-02 paper/paper.tex
finished another review pass

@@
 Sorting has enjoyed an important place within computer programming topics for a long time.
-Apart from having great importance in practical applications, being very useful to organize data for fast retrieval, it also attracts attention as a theoretically interesting problem in itself~\cite{DBLP:books/lib/Knuth98a}.
+It has great importance to practical applications, being very useful to organize data for fast retrieval, and also attracts attention as a theoretically interesting problem in itself~\cite{DBLP:books/lib/Knuth98a}.
 While sorting is a well-understood problem in general, with a few classic algorithms available, achieving optimal performance for specific problems and architectures may require a careful algorithm selection rather than just tuning parameters \cite[Part II, Introduction]{DBLP:books/daglib/0023376}.
@@
 These features include cache memory and paralellism at the processor and instruction levels.
-Cache memory has become popular in consumer processors in the past couple of decades, allowing programs to often attain higher performance than would be possible with a simpler memory architecture, as long as memory access patterns exhibit temporal and spatial locality~\cite[Chapter 3]{Drepper07whatevery}.
+Cache memory has become popular in consumer processors in the past couple of decades, and as long as memory access patterns exhibit temporal and spatial locality, it often allows programs to attain a higher performance than would be possible with a simpler memory architecture~\cite[Chapter 3]{Drepper07whatevery}.
 Writing software that can better cope with parallelism has also become necessary to exploit the full potential offered by modern and future processors~\cite{wilson2018}.
 
-The interest in sorting and the great relevance of cache memory and parallelism in modern computing naturally led to research on sorting algorithms that account for these factors.
+The interest in sorting and the great relevance of cache memory and parallelism to modern computing naturally led to research on sorting algorithms that account for these factors.
 Processors offering SIMD (single instruction, multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
 One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}.
-The current availability of processors with 512 bit registers underscores the need of being able to write software that exploits vector operations, not to mention GPUs.
+The current availability of processors with 512 bit registers underscores the need of being able to write software that exploits vector operations, what can also benefit GPU programming.
 
@@
 %
-\chipsort offers four high-level functions that can be used in applications as replacements for other sorting functions such as \verb sort  in the Julia standard library.
+\chipsort offers four high-level functions that can be used in applications as replacements for other sorting functions such as {\tt sort} in the Julia standard library.
 The {\tt chipsort!} function does not require any configuration and offers a great performance on a wide range of cases.
-The remaining functions are intended to arrays of different sizes, and require choosing some numeric parameters.
+The remaining functions are suitable for arrays of specific size ranges, and require choosing some numeric parameters.
 The function \verb chipsort_large  offers a good overview of all the techniques implemented in the module, and this section is structured according to its stages.
-Some of these techniques might also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.
+Some of these techniques may also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.
 
@@
 5.3.4]{DBLP:books/lib/Knuth98a} are graphs composed by comparator modules, units that take two numbers in, and output them in order: the smallest and the largest at specific outputs.
-Properly arranging the compatators allows the creation of a procedure that takes a vector of numbers in, and outputs them in sequence.
-The design of a sorting network can minimize the total number of comparisons, or also take into account that comparisons can be done in parallel, minimizing the number of steps to complete the process.
+By properly arranging the comparators we can create a procedure to takes a vector of elements and output them in sequence.
+The design of a sorting network can minimize the total number of comparisons, or if they can be done in parallel, minimize the number of steps to complete the process.
 
@@
 \chipsort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}.
-The generated function \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} works by assigning each element of the sequence at each step to a variable, and calculating values from the next step according to the network specification.
+The generated function \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} assigns each element of the sequence at each step to a variable, and the values from the next step are calculated according to the network specification.
 The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types.
@@
 
-The {\tt sort\_net} generated function expects elements with {\tt max} and {\tt min} methods defined.
+The {\tt sort\_net} generated function expects elements to which {\tt max} and {\tt min} are defined.
 It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works.
@@
 Each column in this representation is termed a {\em lane}.
-The result from {\tt sort\_vecs} is that each lane contains a sorted sequence.
+The result from {\tt sort\_net} with vectors is that each lane contains a sorted sequence.
 This data must be transposed to produce sorted vectors.
@@
 Figure~\ref{fig:transpose-vecs} displays an example of 16 random integers loaded into SIMD registers as 4 vectors.
-After {\tt sort\_vecs} each lane contains a sorted sequence, and {\tt transpose\_vecs} turns the lanes into vectors (registers).
-The \verb chipsort_small!
-function is essentially a sorting network followed by a transpose and a merge procedure.
+After {\tt sort\_net} each lane contains a sorted sequence, and {\tt transpose\_vecs} turns the lanes into vectors (registers).
+The {\tt chipsort\_small!} function is essentially a sorting network followed by a transpose and a merge procedure.
 
@@
 Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/bitonic-merge-network.jl#L12}{\tt bitonic\_merge}.
-Merging two sorted vectors requires first reversing one of them, and the result is stored into two new vectors of the same size containing the first and second halves of the total sorted sequence.
+Merging two sorted vectors requires first reversing one of them, and the result is stored in two vectors containing the first and second halves of the complete sorted sequence.
 
@@
 One of the most peculiar aspects of \chipsort, taken directly from AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15}, is the use of the Combsort algorithm.
-Combsort is essentially a generalization of Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{INCERPI198737,dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}.
-The array is swept multiple times, and pairs of values at a given distance from each other are compared and stored in order.
-The distance is reduced at each pass by a factor of $\frac{4}{3}$.
-This algorithm lends itself easily to vectorization, either by parallelizing the operations from a single array or executing the algorithm over independent arrays in parallel.
+Combsort essentially generalizes Bubble sort in the same way that Shell sort generalizes Insertion sort~\cite{INCERPI198737,dobosiewicz1980efficient,Lacey:1991:FES:117187.117218,DBLP:books/lib/Knuth98a,DBLP:books/daglib/0023376}.
+The array is swept multiple times, ordering pairs of values at a given distance from each other.
+The distance is reduced each time by $\frac{4}{3}$.
+Parallelism is achieved either by parallelizing the operations from a single array or by processing independent arrays of the same size.
 
-\chipsort contains two functions that utilize Combsort, employing parallelism in these two distinct forms.
+\chipsort contains two functions that utilize Combsort, employing these two distinct forms of parallelism.
 The first is {\tt chipsort!}, a serial implementation of the algorithm which is vectorized implicitly by the optimizing compiler.
 {\tt chipsort!} starts with Combsort until the interval size is 1, when it switches to Insertion sort.
-This is similar to what is often done in practical implementations of Quicksort, as in the Julia standard library, switching to Insertion sort when the array becomes too small.
-The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort is applied only once through the whole array.
 
@@
 In both implementations switching to Insertion sort presents a better performance than sticking to Combsort until the end.
-It essentially means using Bubble sort once the interval becomes 1, therefore looking for an alternative seems like a good idea.
-Insertion sort should fare well because the array is now approximately sorted, and it should not have to move any out-of-order elements very far from where they are found.
+Once the interval becomes 1 the algorithm essentially turns into Bubble sort, therefore looking for an alternative seems like a good idea.
+Insertion sort should fare well because the array is now approximately sorted, and any out-of-order elements found should be close to their destinations.
 
 Finishing a sorting method with Insertion sort is a common practice, present at least in Musser's Introsort~\cite{musser1997introspective}.
-The utilization of Insertion sort as a final stage to Combsort has been considered before~\cite{combwiki,INCERPI198737}, although the method does not seem to enjoy wide adoption.
-The main contribution of the present work may be to propose it again in the SIMD context, extending AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} and bringing experimental evidence of its good performance.
+This is also akin to the classic method used in the Julia standard library, switching to Insertion sort when the array becomes too small.
+The difference is that Combsort is not a divide-and-conquer approach, and therefore Insertion sort must be applied once on the whole array.
+
+The utilization of Insertion sort as a final stage to Combsort has been considered before~\cite{combwiki,INCERPI198737}, and Insertion sort was even employed by~\cite{DBLP:journals/pvldb/InoueT15} for dealing with partial keys.
+The main contribution of the present work may be to extend AA-sort~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/InoueT15} by giving a larger role to Insertion sort and also relying more on automatic vectorization by the compiler.
 
@@
 In-place matrix transposition can be attained by moving a number to its destination, then moving the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}.
-While carrying out this procedure is simple, it is not trivial to find out what are the cycles to tranpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[Sec.
-1.3.3, Ex.
-12]{DBLP:books/lib/Knuth97}.
+While carrying out this procedure is simple, it is not trivial to find out what are the cycles to transpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[1.3.3-12]{DBLP:books/lib/Knuth97}.
 
 \chipsort contains the generated function \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/transpose-vecs.jl#L19}{\tt transpose!} to perform in-place matrix transposition.
-The cycle seeds are computed at the time of code generation, and at runtime the function only needs to move the data and compute the indices detecting when each cycle finishes.
-The function is therefore completely generic, supporting any dimensions, but with a very simple runtime code, with the most complex part of the problem being tackled programatically during code generation.
+The cycle seeds are computed at the time of code generation, and at run-time the function only moves the data, computing the index sequences and detecting when each cycle finishes.
+Therefore even though the function is completely generic, supporting any dimensions, the most complex part of the problem is tackled programmatically during code generation, resulting in a very simple run-time code.
 
@@
 Given a set of sorted arrays, once the tree is initialized data is taken from it one vector at a time, producing the final sorted sequence.
-Each time a new vector is requested the tree is traversed looking for the smallest element between each of the node children, until one of the input arrays is reached.
-The data from the array is then loaded from memory and merged with the intermediate data, producing new updates moving trought the tree until the root is reached.
+Each time a new vector is requested the tree is traversed following the smallest element among the node children, until one of the input arrays is reached.
+The data from the array is then loaded from memory and merged with the intermediate data, producing new updates moving through the tree until the root is reached.
 
 The intermediate data from the tree is kept by \chipsort in a contiguous array, treated as a heap.
-Assuming 1-based indexing, each node ancestor can be found simply by the Euclidean division of its index by 2.
-During operation it is assumed this data should fit some level of cache memory, as in the related work~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15}.
+Assuming 1-based indexing, each node ancestor can be simply found by the Euclidean division of its index by 2.
+This data is assumed to fit some level of cache memory during operation, as in the related work~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08,DBLP:journals/pvldb/InoueT15}.
 
@@
 \item {\tt insertion\_sort!} Insertion sort taken from the Julia library.
-\item {\tt sort!} the Julia standard sort, Quicksort finishing with Insertion sort for small arrays (20 elements).
+\item {\tt sort!} the Julia standard sort, Quicksort finishing with Insertion sort for small arrays (\href{https://github.com/JuliaLang/julia/blob/de3401908a506727ed70b3e953bd721ab4d66109/base/sort.jl#L461}{20} elements).
 \end{itemize}
@@
 \subsection{Medium and large arrays}
-In our second experiment we compared many different methods on inputs of exponentially increasing sizes, from $2^6$ up to $2^{20}$.
+Our second experiment compared many different methods on inputs of exponentially increasing sizes, from $2^6$ to $2^{20}$.
 Figure~\ref{fig:bench-time} displays the resulting measurements.
-Although the curves look too concentrated in this visualization, it is possible to notice that {\tt chipsort!} consistently outperforms the Julia standard library {\tt sort!} by a factor of 80\% to 100\%, or up to twice as fast.
+Although the curves look too concentrated in this visualization, it is possible to notice that {\tt chipsort!} consistently outperforms the Julia standard library by a factor of 80\% to 100\%, or up to twice as fast.
 We can also observe that Insertion sort soon becomes too inefficient.
@@
 \begin{figure}[htb]
-\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-time.pdf}}
+\centerline{\includegraphics[width=\linewidth]{fig/chipsort-bench-time.pdf}}
 \caption{Running time of the different sorting methods studied.}
@@
 
-Figure~\ref{fig:bench-curves} shows a different visualization from the same data in Figure~\ref{fig:bench-time}, with the times divided by the input size at each point.
-The vertical dashed lines indicate where the input array would be as big as each cache level.
+Figure~\ref{fig:bench-curves} shows a different visualization from the same data with the time divided by the input size.
+The vertical dotted lines indicate the size of each cache level.
 Here we can see more clearly how {\tt chipsort\_medium!} and {\tt chipsort\_large} slightly outperform {\tt chipsort!} in a few cases.
 
-\begin{figure*}[thb]
-\centerline{\includegraphics[width=0.7\linewidth]{fig/chipsort-bench-curves.pdf}}
-\caption{Curves displaying the relative running time from each studied method. The ordinate represents the median of the measured running times divided by the input size at each test.}
+\begin{figure}[htb]
+\centerline{\includegraphics[width=\linewidth]{fig/chipsort-bench-curves.pdf}}
+\caption{Curves displaying the relative running time from each studied method.
+The ordinate represents the median of the measured running times divided by the input size at each test.
+Lower is faster.}
 \label{fig:bench-curves}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure}[htb]
-\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-8k.pdf}}
-\centerline{\includegraphics[width=0.9\linewidth]{fig/chipsort-bench-256k.pdf}}
+\begin{figure*}[htb]
+  \centerline{\includegraphics[width=0.4\linewidth]{fig/chipsort-bench-8k.pdf}
+    \includegraphics[width=0.4\linewidth]{fig/chipsort-bench-256k.pdf}}
 \caption{Empirical CDF from different methods with 8k and 256k {\tt UInt32} values. Results from 10,000 measurements.}
 \label{fig:bench-cdf}
-\end{figure}
+\end{figure*}
 
@@
 Results obtained for C++ with Julia inter-operation using Cxx.jl were slightly worse than native, thus discarded.
-While Numpy was almost as fast as Julia, no reason could be found for the dismal performance from the C++ standard library.
+While Numpy was almost as fast as Julia, no reason could be found for the dismal performance of the C++ standard library.
 
@@
 \tabcolsep11pt
-\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different platforms.}{
+\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different software platforms.}{
 \begin{tabular}{|l|r|r|}\hline
@@
 \tabcolsep11pt
-\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different platforms.}{
-\begin{tabular}{|l|r|r|}\hline
-Method & 8k & 1M\\\hline
-AXV512 \chipsort & 177.4 $\mu$s & 35.20 ms\\\hline
-AXV512 Julia standard & 347.5 $\mu$s & 69.07 ms\\\hline
-NEON \chipsort & 350.9 $\mu$s & 73.36 ms\\\hline
-NEON Julia standard & 573.2 $\mu$s & 113.68 ms\\\hline
+\tbl{Sorting arrays of 8k and 1M {\tt UInt32} elements in different hardware platforms.}{
+\begin{tabular}{|l|l|r|r|}\hline
+Chip & Method & 8k & 1M\\\hline
+AXV512 & \chipsort & 177.4 $\mu$s & 35.20 ms\\\hline
+AXV512 & Julia standard & 347.5 $\mu$s & 69.07 ms\\\hline
+NEON & \chipsort & 350.9 $\mu$s & 73.36 ms\\\hline
+NEON & Julia standard & 573.2 $\mu$s & 113.68 ms\\\hline
 \end{tabular}}
@@
 No change in performance can be seen in the AVX512 machine, and on the ARM machine the speedup was smaller than before: only 60\%.
-The most notable fact in this experiment is the fact the same code was successfully employed to produce a vectorized sort on machines with different vector sizes and even instruction sets, always with beneficial effects.
+The most notable fact in this experiment is that the same high-level and serial code successfully produced a vectorized sort on machines with different vector sizes and even instruction sets, always with a beneficial outcome.
 
@@
 This method was implemented as a serial program vectorized by the Julia compiler and LLVM, resulting in a sorting function up to twice as fast as the Julia standard library sort.
-This performance holds for at least a million 32-bit integers in a conventional computer.
+This performance holds for at least a million 32-bit integers in a personal computer.
 
@@
 This contrasts with other libraries where different networks are implemented straight as final functions.
-Julia also has the opportunity to perform many optimizations since \chipsort is implemented in pure Julia with abstract methods that can be specialized.
+Julia also has the opportunity to perform custom optimizations since \chipsort is implemented in pure Julia with abstract methods that can be specialized.
 
@@
 It is just a simple, scalar implementation of Combsort that is optimized by the compiler to generate code adapted to different problem settings and to the hardware architecture at run-time.
-And in the cases where explicit vectorization can produce better code Julia still offers a suitable framework that allowed \chipsort to implement sophisticated techniques with concise code.
+And in the cases where explicit vectorization is necessary Julia still offers a suitable framework that allowed \chipsort to implement sophisticated techniques with concise code.
 This project can illustrate how well Julia implements traditional language features such as parametric types, and its metaprogramming features make it stand out from other languages even more.
 
-Some future prospects for \chipsort are to better deal with floating-point comparisons; implement stable sorting of key-payload data; explore the limits of in-register processing, specially in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to produce optimized code for complex tasks such as multi-stage multi-threaded merge sorting of large arrays.
+Some future prospects for \chipsort are to implement missing features such as stable sorting of key-payload data and buffers in the merge tree; explore the limits of in-register processing in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to optimize code for complex tasks such as multi-stage multi-threaded merge-sorting of large arrays.
 

### 2019-06-02 paper/paper.tex
jl

@@
 It can be called with simple data types, {\em e.g.} {\tt sort\_net(4,2,5,3)} works.
-In practice \chipsort often calls this function with vectors, completely relying on SIMD.jl~\cite{erik_schnetter_2019_2592633}.
+In practice \chipsort often calls this function with vectors, completely relying on {\tt SIMD.jl}~\cite{erik_schnetter_2019_2592633}.
 This module maps a vector in Julia similar to tuples or static arrays straight to LLVM vector types, ensuring SIMD code can be produced when possible and also allowing the flexibility of handling logic vectors that are larger than the actual registers, or spilling data to the stack if all registers are occupied.
@@
 For the case of Python the measurements obtained by calling the function through PyCall were better than using the native Python shell, therefore the best value was kept.
-Results obtained for C++ with Julia inter-operation using Cxx.jl were slightly worse than native, thus discarded.
+Results obtained for C++ with Julia inter-operation using {\tt Cxx.jl} were slightly worse than native, thus discarded.
 While Numpy was almost as fast as Julia, no reason could be found for the dismal performance of the C++ standard library.
@@
 
-While the main \chipsort method relies on compiler optimizations, other methods utilize explicit vectorization trough SIMD.jl.
+While the main \chipsort method relies on compiler optimizations, other methods utilize explicit vectorization trough {\tt SIMD.jl}.
 \chipsort makes extensive use of metaprogramming, more specifically Julia generated functions.

### 2019-06-02 paper/paper.tex
jl2
jll

@@
 While there are well-known sorting algorithms for conventional computers, special techniques are required to take the most out of real hardware offering features such as cache memory and instruction-level parallelism.
-ChipSort is a Julia module for SIMD and cache-aware sorting.
+ChipSort.jl is a Julia module for SIMD and cache-aware sorting.
 It implements sorting networks and bitonic merge networks with SIMD instructions, with configurable vector sizes.
@@
 
-This article presents \chipsort, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
+This article presents {\tt ChipSort.jl}, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
 These include sorting and merging networks, the use of Combsort for medium sized arrays and of multi-way merging for large arrays.

### 2019-06-02 paper/paper.tex
tt

@@
 The remaining functions are suitable for arrays of specific size ranges, and require choosing some numeric parameters.
-The function \verb chipsort_large  offers a good overview of all the techniques implemented in the module, and this section is structured according to its stages.
+The function {\tt chipsort\_large} offers a good overview of all the techniques implemented in the module, and this section is structured according to its stages.
 Some of these techniques may also be useful in other applications instead of sorting, one clear example being the in-place matrix transpose.
@@
 
-The function \verb chipsort_large  achieves sorting by first splitting the input array in chunks that are sorted separately with \verb chipsort_medium!
-and then merged simultaneously.
+The function {\tt chipsort\_large} achieves sorting by first splitting the input array in chunks that are sorted separately with {\tt chipsort\_medium!}  and then merged simultaneously.
 This inner sort is performed in multiple steps:
@@
 Other forms of instruction-level parallelism can also be in play, though.
-It may be possible to parallelize instructions such as \verb min  and \verb max , or the loading of data into registers.
+It may be possible to parallelize instructions such as {\tt min} and {\tt max}, or the loading of data into registers.
 This is often done implicitly by the microarchitecture, and programmers can only make sure their code is suitable for this by properly ordering operations and avoiding branching, for instance.
@@
 
-\chipsort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}.
-The generated function \href{https://github.com/nlw0/\chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} assigns each element of the sequence at each step to a variable, and the values from the next step are calculated according to the network specification.
+\chipsort supports sorting networks of different sizes, currently only powers of 2, which are predefined as data structures in the file \href{https://github.com/nlw0/chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-network-parameters.jl}{\tt sorting-network-parameters.jl}.
+The generated function \href{https://github.com/nlw0/chipsort.jl/blob/d2d049b7413f0073476021fa62fb748803130768/src/sorting-networks.jl#L13}{\tt sort\_net} assigns each element of the sequence at each step to a variable, and the values from the next step are calculated according to the network specification.
 The comparisons are performed by the {\tt min} and {\tt max} functions, and the function is generic on the element types.

### 2019-06-04 paper/paper.tex
few changes prior to submission

@@
 It has great importance to practical applications, being very useful to organize data for fast retrieval, and also attracts attention as a theoretically interesting problem in itself~\cite{DBLP:books/lib/Knuth98a}.
-While sorting is a well-understood problem in general, with a few classic algorithms available, achieving optimal performance for specific problems and architectures may require a careful algorithm selection rather than just tuning parameters \cite[Part II, Introduction]{DBLP:books/daglib/0023376}.
+While sorting is a well-understood problem in general, with a few classic algorithms available, achieving optimal performance for specific problems and architectures may require a careful algorithm selection rather than a simple parameter tuning \cite[Part II, Introduction]{DBLP:books/daglib/0023376}.
 
-The running time of a computer program is governed by many factors, starting with algorithmic complexity and the base clock speed of the processor.
-The performance observed in real computers has been increasingly depending in processor features that are not taken into account by the simplest models, though.
-These features include cache memory and paralellism at the processor and instruction levels.
-Cache memory has become popular in consumer processors in the past couple of decades, and as long as memory access patterns exhibit temporal and spatial locality, it often allows programs to attain a higher performance than would be possible with a simpler memory architecture~\cite[Chapter 3]{Drepper07whatevery}.
-Writing software that can better cope with parallelism has also become necessary to exploit the full potential offered by modern and future processors~\cite{wilson2018}.
+Many factors can influence the running time of a computer program, starting with algorithmic complexity and the base clock speed of the processor.
+The performance observed in real computers has been increasingly depending on processor features that are not taken into account by the simplest models, though.
+These features include cache memory and parallelism at the processor and instruction levels.
+Cache memory has become popular in consumer processors in the past couple of decades, and as long as memory access patterns exhibit temporal and spatial locality, it allows programs to attain a higher performance than would be possible with a simpler memory architecture~\cite[Chapter 3]{Drepper07whatevery}.
+Writing software with better support for parallelism has also become necessary to utilize the full potential offered by modern and future processors~\cite{wilson2018}.
 
-The interest in sorting and the great relevance of cache memory and parallelism to modern computing naturally led to research on sorting algorithms that account for these factors.
+The interest in sorting and the great relevance of cache memory and parallelism to modern computing have naturally led to research on sorting algorithms that account for these factors.
 Processors offering SIMD (single instruction, multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
 One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}.
-The current availability of processors with 512 bit registers underscores the need of being able to write software that exploits vector operations, what can also benefit GPU programming.
+The current availability of processors with 512 bit registers underscores the necessity of being able to write software that utilizes vector operations, what can also benefit GPU programming.
 
@@
 This article presents {\tt ChipSort.jl}, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
-These include sorting and merging networks, the use of Combsort for medium sized arrays and of multi-way merging for large arrays.
+These include sorting and merging networks, the use of Combsort with Insertion sort finalization for medium sized arrays and of multi-way merging for large arrays.
 These techniques require not only tuning parameters such as vector and array sizes, but also generating specific code when implementing different size networks.
@@
 \item sorting small blocks of data with a sorting network;
-\item reordering vectors in-place according to a matrix transpose;
+\item reordering vectors in memory according to a matrix transpose;
 \item vectorized Combsort with a limited number of iterations;
@@
 Sorting networks~\cite[Sec.
-5.3.4]{DBLP:books/lib/Knuth98a} are graphs composed by comparator modules, units that take two numbers in, and output them in order: the smallest and the largest at specific outputs.
-By properly arranging the comparators we can create a procedure to takes a vector of elements and output them in sequence.
+5.3.4]{DBLP:books/lib/Knuth98a} are graphs composed by comparator modules, units that take two numbers in, and output the smallest and largest at specific outputs.
+By properly arranging comparators we can create a procedure to take an array of elements and output them in sequence.
 The design of a sorting network can minimize the total number of comparisons, or if they can be done in parallel, minimize the number of steps to complete the process.
@@
 A bitonic sequence starts either increasing or decreasing and contains at most one change of direction.
-Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort in a single generated function, \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/bitonic-merge-network.jl#L12}{\tt bitonic\_merge}.
+Bitonic merge networks allow the creation of merge networks of different sizes, and they are again implemented in \chipsort with a single generated function, \href{https://github.com/nlw0/ChipSort.jl/blob/10eb1a962e720aee2bf8733c832468b51a782c1a/src/bitonic-merge-network.jl#L12}{\tt bitonic\_merge}.
 Merging two sorted vectors requires first reversing one of them, and the result is stored in two vectors containing the first and second halves of the complete sorted sequence.
@@
 Apart from the use in sorting small sequences and in the multi-way merge tree, bitonic merge networks can also be used to implement a regular merge sort.
-This technique was investigated in \chipsort only superficially, and while there were no promising results in terms of performance, \chipsort also contains the generated function {\tt bitonic\_merge\_interleaved} to illustrate how metaprogramming can help implementing multiple simultaneous merges with interleaved execution, as done in other projects~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
+This technique was investigated in \chipsort only superficially, and while there were no promising results in terms of performance, \chipsort also contains the generated function {\tt bitonic\_merge\_interleaved} to illustrate how metaprogramming can help implement multiple simultaneous merges with interleaved execution, as done in other projects~\cite{DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
 
@@
 
-After the first two steps the result is essentially a matrix where the $K\times J$ columns are the vectors, and each row contains an approximately sorted sequence.
+After the first two steps the result is essentially a matrix where the $K\times J$ columns are the vectors, and each of the $V$ rows contains an approximately sorted sequence.
 The transpose steps reshape each row into a new block.
@@
 In both implementations switching to Insertion sort presents a better performance than sticking to Combsort until the end.
-Once the interval becomes 1 the algorithm essentially turns into Bubble sort, therefore looking for an alternative seems like a good idea.
-Insertion sort should fare well because the array is now approximately sorted, and any out-of-order elements found should be close to their destinations.
+Once the interval becomes 1 Combsort essentially becomes Bubble sort, thus finding an alternative seems beneficial.
+Insertion sort should fare well because the array is now approximately sorted, and any out-of-order elements should be already close to their destinations.
 
@@
 %
-In-place matrix transposition can be attained by moving a number to its destination, then moving the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}.
+In-place matrix transposition can be attained by moving a number to its destination, then moving away the number that was found there, and so on, until a cycle is completed and a new number is selected to be moved~\cite{10.1093/comjnl/2.1.47}.
 While carrying out this procedure is simple, it is not trivial to find out what are the cycles to transpose a matrix of any given dimensions, and even finding out the number of cycles turns out to be a difficult problem~\cite[1.3.3-12]{DBLP:books/lib/Knuth97}.
@@
 The cycle seeds are computed at the time of code generation, and at run-time the function only moves the data, computing the index sequences and detecting when each cycle finishes.
-Therefore even though the function is completely generic, supporting any dimensions, the most complex part of the problem is tackled programmatically during code generation, resulting in a very simple run-time code.
+Therefore even though the function is completely generic, supporting any shape, the most complex part of the problem is tackled programmatically during code generation, resulting in a very simple run-time code.
 
@@
 The term {\it method} can be understood as both the algorithms and their implementations in the Julia language.
-The specific methods tested are:
+The specific methods tested were:
 \begin{itemize}
@@
 The {\tt chipsort\_small!} method based on sorting and bitonic merge networks was compared to serial Insertion sort, the Julia sort and {\tt chipsort!}.
-The small running time makes it challenging to perform proper measurements in this experiment.
-Our task was therefore to sort 128 sequences of 64 contiguous values in memory, forming a single block of $2^{13}$ uniform random entries.
-Sorting was then performed in-place, in a for-loop.
+The short running time makes it challenging to perform proper measurements, therefore the task was to sort 128 sequences of 64 contiguous values in memory, forming a single block of $2^{13}$ uniform random entries.
+Sorting was performed in-place, in a for-loop.
 The results are in Table~\ref{tab:bench-small}.
@@
 Method & Median time\\\hline
-\chipsort small & 57.206 $\mu$s\\\hline
-\chipsort & 133.510 $\mu$s\\\hline
+{\tt chipsort\_small!} & 57.206 $\mu$s\\\hline
+{\tt chipsort!} & 133.510 $\mu$s\\\hline
 Insertion sort & 194.679 $\mu$s\\\hline
@@
 
-\subsection{Other languages}
-A simple experiment was also carried out with different programming languages in the same machine, and with Julia in different machines to provide a richer context.
+\subsection{Other platforms}
+A final experiment was carried out with different programming languages in the same machine and with Julia in different machines, to provide more context.
 Table~\ref{tab:bench-cxx} displays statistics collected from Julia, Python and C++ programs.
@@
 
-Table~\ref{tab:bench-512} contains the results for {\tt chipsort!} and the Julia standard {\tt sort!} in an Intel processor with AVX512 instructions, and an ARM processor with NEON instructions.
+Table~\ref{tab:bench-512} contains the results for {\tt chipsort!} and the Julia standard {\tt sort!} on an Intel processor with AVX512 instructions, and an ARM processor with 128-bit NEON instructions.
 \begin{table}[h]
@@
 \end{table}
-
-No change in performance can be seen in the AVX512 machine, and on the ARM machine the speedup was smaller than before: only 60\%.
+No change in performance can be seen on the AVX512 machine, and on the ARM machine the speedup was smaller than before: only 60\%.
 The most notable fact in this experiment is that the same high-level and serial code successfully produced a vectorized sort on machines with different vector sizes and even instruction sets, always with a beneficial outcome.
@@
 
-Some future prospects for \chipsort are to implement missing features such as stable sorting of key-payload data and buffers in the merge tree; explore the limits of in-register processing in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to optimize code for complex tasks such as multi-stage multi-threaded merge-sorting of large arrays.
+Some future prospects for \chipsort are to implement missing features such as buffers in the merge tree and stable sorting of key-payload data; explore the limits of in-register processing in modern 512-bit architectures an eventually GPUs; explore how to interact with the compiler in order to optimize code for complex tasks such as multi-stage multi-threaded merge-sorting of large arrays.
 

### 2019-06-22 paper/paper.tex
versions, and gcc re-run

@@
 \subsection{Other platforms}
-A final experiment was carried out with different programming languages in the same machine and with Julia in different machines, to provide more context.
+One last experiment was carried out with different programming languages in the same machine and with Julia in different machines, to provide more context.
 Table~\ref{tab:bench-cxx} displays statistics collected from Julia, Python and C++ programs.
+Versions utilized were Julia 1.3.0-DEV.431, Python 3.6.8 with Numpy 1.14.3 and GCC 8.3.0 with libstdc++ 6.0.25 on a Ubuntu 18.04 system.
 For the case of Python the measurements obtained by calling the function through PyCall were better than using the native Python shell, therefore the best value was kept.
@@
 \chipsort & 172.35 $\mu$s & 35.28 ms\\\hline
-Julia standard & 345.46 $\mu$s & 68.39 ms\\\hline
-Numpy via PyCall & 369.50 $\mu$s & 70.01 ms\\\hline
-C++ {\tt std::qsort} & 788.59 $\mu$s & 147.37 ms\\\hline
-C++ {\tt std::sort} & 1,207.23 $\mu$s & 224.71 ms\\\hline
+Julia standard & 337.03 $\mu$s & 68.39 ms\\\hline
+Numpy via PyCall & 354.55 $\mu$s & 70.01 ms\\\hline
+C++ {\tt std::qsort} & 770.93 $\mu$s & 143.48 ms\\\hline
+C++ {\tt std::sort} & 1,177.54 $\mu$s & 221.27 ms\\\hline
 \end{tabular}}

### 2019-06-22 paper/paper.tex
simd info

@@
 
+One particular form of parallelism that concerns the present work are SIMD (single instruction, multiple data) instructions~\cite{SIMDtut}.
+This is a form of computation where a same operation is applied to different independent memory elements simultaneously.
+These operations can be described as operating over {\em vectors}, where each {\em lane} contains the same data that would be involved in carrying out the same operations as a scalar version of the same program.
+SIMD parallelism is only suitable to specific problems, and its use may imply in memory access constraints and even clock speed reductions.
+There are still cases where it can bring great performance while being conveniently employed.
+
 The interest in sorting and the great relevance of cache memory and parallelism to modern computing have naturally led to research on sorting algorithms that account for these factors.
-Processors offering SIMD (single instruction, multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
+Processors offering SIMD instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
 One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}.

### 2019-06-22 paper/paper.tex
moar simd and metaprogramming

--- 2019-06-22 paper/paper.tex
@@
 
-The interest in sorting and the great relevance of cache memory and parallelism to modern computing have naturally led to research on sorting algorithms that account for these factors.
-Processors offering SIMD (single instruction, multiple data) instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
+One particular form of parallelism that concerns our work is SIMD (single instruction, multiple data) instructions, where the processor applies a same operation to multiple independent memory elements simultaneously.
+These operations are said to act on {\em vectors}, where each parallel {\em lane} carries out the scalar version of the program.
+SIMD parallelism is only suitable to specific problems.
+It may introduce data access constraints and overheads, and even clock speed reductions.
+SIMD performance can be affected negatively by branching and scattered memory access.
+It is mostly suited for calculating many replicas of a long arithmetic expression, preferably with little IO.
+It can still pay off in appropriate settings and, as will be demonstrated, sometimes the implementation can be simple too.
+More information about SIMD can be found in~\cite{SIMDtut}.
+
+Processors offering SIMD instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
 One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}.
-The current availability of processors with 512 bit registers underscores the necessity of being able to write software that utilizes vector operations, what can also benefit GPU programming.
+The current availability of processors with 512 bit registers, along with modern GPUs, underscore the demand for software-building tools that support vector operations.
+One programming language feature relevant to that goal is polymorphism, which allows programmers to abstract over data types and vectorization.
+Another is {\em metaprogramming}, the ability of writing programs that manipulate or generate other programs~\cite{Abelson1996}\cite[Ch.25]{JuliaDocs}.
+Metaprogramming enables code generation for complex algorithms customized to problem parameters and platform resources.
 
+The interest in sorting and the great relevance of cache memory and parallelism to modern computing eventually led to research on sorting algorithms that account for these factors, following research in other topics where SIMD is more clearly beneficial.
 At the time SIMD CPUs became widely available there was already a trend to employ specialized hardware like GPUs for scientific applications~\cite{larsen2001fast,DBLP:conf/micro/ThompsonHO02}.

### 2019-06-22 paper/paper.tex
note on uint32 and floats

@@
 
-Processors offering SIMD instructions and other forms of instruction-level parallelism started to become widely available in the early 2000s.
+Processors offering SIMD or other forms of instruction-level parallelism started to become widely available in the early 2000s.
 One example is the introduction of SSE instructions in the x86 architecture~\cite[Chapter 1]{kusswurm18}.
 The current availability of processors with 512 bit registers, along with modern GPUs, underscore the demand for software-building tools that support vector operations.
-One programming language feature relevant to that goal is polymorphism, which allows programmers to abstract over data types and vectorization.
+One programming language feature relevant to that goal is polymorphism, allowing programmers to abstract over data types and vectors.
 Another is {\em metaprogramming}, the ability of writing programs that manipulate or generate other programs~\cite{Abelson1996}\cite[Ch.25]{JuliaDocs}.
@@
 Although SIMD was used mostly for multimedia applications at first~\cite{CHEN2006509,DBLP:journals/mam/SlingerlandS05}, some early alternative uses include 3D graphics~\cite{DBLP:conf/pcm/MaY02} and machine learning~\cite{DBLP:conf/europar/StreyB01}.
-
 Following earlier work on parallel sorting in other architectures, the late 2000s saw the first practical demonstrations of sorting with modern consumer SIMD chips~\cite{DBLP:conf/IEEEpact/InoueMKN07,DBLP:journals/pvldb/ChhuganiNLMHCBKD08}.
@@
 
-The way \chipsort  works mostly follows~\cite{DBLP:journals/pvldb/InoueT15}, although \chipsort does not support yet registers with key and payload.
-Another main difference is that Combsort is always employed with an Insertion sort finalization.
+The way \chipsort works mostly follows~\cite{DBLP:journals/pvldb/InoueT15}, except it does not yet support registers with key and payload and there are no measures to guarantee a stable sort, such as ensuring unique keys or performing a post-processing to enforce the original order of colliding keys.
+Another main difference is that \chipsort always employs Combsort with an Insertion sort finalization.
 
@@
 
+While our experiments focused on 32-bit integer values, similar results were observed with other types, although with possibly smaller speedups with 64-bits.
+Further details are omitted for the sake of brevity.
+For floating-point numbers, while in theory every method should perform comparably to integers, issues can appear in practice due to peculiarities such as {\tt NaN}s.
+This interferes with the ability of the compiler to optimize {\tt chipsort!}, although the other methods with explicit vectorization are less affected.
+There is ongoing work in Julia and LLVM to improve the performance of floating-point comparisons~\cite{llvmD62272,llvm3acc4236b871,Julia31442}.
+
 \subsection{Small arrays}

### 2019-06-22 paper/paper.tex
chipsort github reference

@@
 
-This article presents {\tt ChipSort.jl}, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
+This article presents {\tt ChipSort.jl}~\cite{nicolau_leal_werneck_2019_3228848}, a Julia module that implements SIMD and cache-aware sorting utilizing some of the techniques found in the literature.
 These include sorting and merging networks, the use of Combsort with Insertion sort finalization for medium sized arrays and of multi-way merging for large arrays.

### 2019-06-23 paper/paper.tex
gcc -O3

@@
 Table~\ref{tab:bench-cxx} displays statistics collected from Julia, Python and C++ programs.
-Versions utilized were Julia 1.3.0-DEV.431, Python 3.6.8 with Numpy 1.14.3 and GCC 8.3.0 with libstdc++ 6.0.25 on a Ubuntu 18.04 system.
+The experiment was performed with a Ubuntu 18.04 system using Julia 1.3.0-DEV.431, Python 3.6.8 with Numpy 1.14.3 and GCC 8.3.0 with {\tt -O3 -march=native} and libstdc++ 6.0.25.
 For the case of Python the measurements obtained by calling the function through PyCall were better than using the native Python shell, therefore the best value was kept.
 Results obtained for C++ with Julia inter-operation using {\tt Cxx.jl} were slightly worse than native, thus discarded.
-While Numpy was almost as fast as Julia, no reason could be found for the dismal performance of the C++ standard library.
+Numpy, C++ {\tt sort} and Julia all presented similar performance, unmatched by C++ {\tt qsort}.
+These methods make little to no use of SIMD, enabling \chipsort to surpass them.
 
@@
 Numpy via PyCall & 354.55 $\mu$s & 70.01 ms\\\hline
-C++ {\tt std::qsort} & 770.93 $\mu$s & 143.48 ms\\\hline
-C++ {\tt std::sort} & 1,177.54 $\mu$s & 221.27 ms\\\hline
+C++ {\tt std::sort} & 374.20 $\mu$s & 68.10 ms\\\hline
+C++ {\tt std::qsort} & 589.52 $\mu$s & 108.2 ms\\\hline
 \end{tabular}}
